\documentclass[11pt,a4paper]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}

\definecolor{linkblue}{HTML}{1a5276}
\hypersetup{colorlinks=true,urlcolor=linkblue,linkcolor=linkblue}

\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[C]{\small\thepage}

\titleformat{\section}{\large\bfseries\scshape}{}{0em}{}[\titlerule]
\titlespacing*{\section}{0pt}{12pt}{6pt}

\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}

\begin{document}

\begin{center}
{\LARGE\bfseries Ya\'e Ulrich Gaba}\\[4pt]
{\normalsize Research $\cdot$ Teaching $\cdot$ Consulting}\\[4pt]
{\small Kigali, Rwanda \quad $|$ \quad \href{mailto:gabayae2@gmail.com}{gabayae2@gmail.com} \quad $|$ \quad \href{https://gabayae.github.io}{gabayae.github.io}}\\[2pt]
{\small \href{https://github.com/gabayae}{GitHub: gabayae} \quad $|$ \quad \href{https://scholar.google.com/citations?user=UTszjV4AAAAJ}{Google Scholar} \quad $|$ \quad \href{https://orcid.org/0000-0001-8128-9704}{ORCID: 0000-0001-8128-9704}}
\end{center}

\vspace{4pt}

\section{Summary}

Topologist and AI researcher. I develop mathematical methods for machine learning, teach them at the graduate level, and consult with organizations that need rigorous thinking applied to hard problems. Head of R\&D at a technology startup, leading applied research with cross-industry experience in banking, energy, insurance, IT, and retail.

\section{Education}

\textbf{PhD in Topology} \hfill 2013--2017\\
University of Cape Town (UCT), Cape Town, South Africa\\
{\small Thesis: Asymmetric topology and fixed point theory in generalized metric spaces}

\medskip

\textbf{MSc in Pure \& Applied Mathematics} \hfill 2011--2013\\
African University of Science and Technology (AUST), Abuja, Nigeria

\section{Experience}

\textbf{Head of R\&D $\cdot$ AI Research \& Innovation Lead} \hfill 2025--present\\
AIRINA Labs, Kigali, Rwanda\\
{\small Applied research bridging methodological innovation and operational deployment. Statistical modeling and ML across banking, energy, insurance, IT, and retail. AI research for the African continent.}

\medskip

\textbf{Research Affiliate} \hfill 2020--present\\
African Centre for Advanced Studies (ACAS), Yaound\'e, Cameroon\\
{\small Research in topology, data science, and machine learning. Mentoring African graduate students.}

\medskip

\textbf{Research Associate} \hfill 2021--2024\\
Quantum Leap Africa (QLA) | AIMS-RIC, Kigali, Rwanda\\
{\small Topological methods in machine learning and data analysis. WoComToQC workshop organizer.}

\medskip

\textbf{Postdoctoral Researcher} \hfill 2017--2019\\
North-West University, South Africa\\
{\small Fixed point theory and generalized metric spaces. Published in Topology and its Applications, JMAA, and Quaestiones Mathematicae.}

\medskip

\textbf{Assistant Professor} \hfill 2018--2020\\
IMSP (Institut de Math\'ematiques et de Sciences Physiques), Dangbo, B\'enin\\
{\small Teaching graduate courses in topology and analysis. Supervising MSc students.}

\section{Selected Publications}

\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
\item C.~M.~Farrelly and Y.~U.~Gaba, \emph{The Shape of Data: Geometry-Based Machine Learning and Data Analysis in R}, No Starch Press, 2024.
\item D.~K.~Kadurha, D.~J.~L.~Moutouo, and Y.~U.~Gaba, ``Bellman Operator Convergence Enhancements in Reinforcement Learning Algorithms,'' arXiv:2505.14564, 2025.
\item D.~K.~Kadurha and Y.~U.~Gaba, ``Topological Foundations of Reinforcement Learning,'' arXiv:2410.03706, 2024.
\item Y.~U.~Gaba and E.~Karap\i nar, ``Interpolative Kannan Contractions in $T_0$-Quasi-Metric Spaces,'' \emph{Journal of Mathematics}, 2021.
\item Y.~U.~Gaba, E.~Karap\i nar, A.~Petru\c{s}el, and S.~Radenovi\'c, ``New Results on Start-Points for Multi-Valued Maps,'' \emph{Axioms}, 9(4):141, 2020.
\item Y.~U.~Gaba and E.~Karap\i nar, ``A New Approach to the Interpolative Contractions,'' \emph{Axioms}, 8(4):110, 2019.
\item Y.~U.~Gaba, ``Startpoints and ($\alpha$, $\gamma$)-Contractions in Quasi-Pseudometric Spaces,'' \emph{Journal of Mathematics}, 2014.
\end{itemize}

\section{Skills}

\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
\item \textbf{Mathematics:} Asymmetric Topology, Fixed Point Theory, Generalized Metric Spaces, TDA, Persistent Homology
\item \textbf{Machine Learning \& AI:} Geometric Deep Learning, Deep RL, Graph Neural Networks, Manifold Learning
\item \textbf{Statistical Modeling:} Regression, Time Series, Bayesian Methods, Model Validation, Risk Modeling
\item \textbf{Programming:} Python, R, PyTorch, TensorFlow, scikit-learn, GUDHI, Ripser, SQL
\item \textbf{Tools:} Git, Docker, \LaTeX, Linux, Jupyter, MLflow, GitHub Actions
\item \textbf{Domains:} Banking, Energy, Insurance, IT, Retail, Quantitative Finance
\end{itemize}

\section{Languages}

French (native) $\cdot$ English (fluent)

\section{Community \& Service}

\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
\item Workshop on Computational Topology \& Quantum Computing (WoComToQC) --- organizer
\item Data Science Makers / AI.Technipreneurs --- open-source ML teaching and outreach
\item Data Science Africa --- instructor and participant
\end{itemize}

\end{document}
