% ============================================================
% MATHEMATICAL STATISTICS — Course Notes (L2/L3)
% English version
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,decorations.pathreplacing,shapes}

% --- Python code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true,python3=true}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames]{xcolor}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}
\usepackage{cleveref}

% --- Bibliography ---
\usepackage[numbers]{natbib}

% --- Index ---
\usepackage{makeidx}
\makeindex

% ============================================================
% Theorem environments
% ============================================================
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}[definition]{Example}
\newtheorem{exercise}{Exercise}[chapter]

\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}

\theoremstyle{remark}
\newtheorem{remark}[definition]{Remark}

% ============================================================
% Colored boxes
% ============================================================
\newtcolorbox{keyformulas}[1][]{%
  colback=blue!5,colframe=NavyBlue,title=\textbf{Key Formulas},
  fonttitle=\bfseries,breakable,#1}

\newtcolorbox{warning}[1][]{%
  colback=red!5,colframe=BrickRed,title=\textbf{Common Pitfall},
  fonttitle=\bfseries,breakable,#1}

\newtcolorbox{intuition}[1][]{%
  colback=green!5,colframe=ForestGreen,title=\textbf{Statistical Intuition},
  fonttitle=\bfseries,breakable,#1}

\newtcolorbox{python}[1][]{%
  colback=gray!5,colframe=gray!80,title=\textbf{Python Implementation},
  fonttitle=\bfseries,breakable,#1}

% ============================================================
% Custom commands
% ============================================================
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\iid}{\overset{\text{i.i.d.}}{\sim}}
\newcommand{\xbar}{\bar{X}}
\newcommand{\phat}{\hat{p}}
\newcommand{\thetahat}{\hat{\theta}}
\DeclareMathOperator{\MSE}{MSE}
\DeclareMathOperator{\Bias}{Bias}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

% ============================================================
\begin{document}

\frontmatter
\begin{titlepage}
\begin{tikzpicture}[remember picture, overlay]
  \fill[blue!10!white] (current page.south west) rectangle (current page.north east);
  \fill[blue!80!black, rotate around={-15:(current page.center)}]
    ([xshift=-4cm,yshift=-1cm]current page.center) rectangle ([xshift=4cm,yshift=1.5cm]current page.center);
  \node[anchor=center, text=white, font=\Huge\bfseries, rotate=-15]
    at ([yshift=0.3cm]current page.center) {Statistics};
  \node[anchor=center, text=orange!80!yellow, font=\Large, rotate=-15]
    at ([yshift=-1cm]current page.center) {Lecture Notes};
  \node[anchor=north, text=blue!70!black, font=\Large\itshape]
    at ([yshift=-4cm]current page.center) {Ya\'e Ulrich Gaba};
  \node[anchor=north, text=blue!50!black, font=\large]
    at ([yshift=-5.5cm]current page.center) {Licence L3 --- 2025--2026};
  \fill[orange!70!yellow] ([yshift=3cm]current page.south west) rectangle ([yshift=3.3cm]current page.south east);
  \node[anchor=south, text=gray, font=\small] at ([yshift=1cm]current page.south) {\today};
\end{tikzpicture}
\end{titlepage}
\tableofcontents

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Mathematical statistics is the science of drawing rigorous conclusions from uncertain data. Across medicine, physics, economics, and artificial intelligence, decisions rely on data analysis. This course provides the theoretical and practical foundations of the discipline.

\medskip
\textbf{Prerequisites.} Probability theory (random variables, classical distributions, convergence), real analysis~I (series, integrals), linear algebra (vector spaces, matrices).

\medskip
\textbf{Organization.} Each chapter opens with a motivating example, develops theory rigorously, includes Python implementations, and concludes with graded exercises:
\begin{itemize}
\item[\textbf{$\star$}] Computational exercises
\item[\textbf{$\star\star$}] Theoretical exercises
\item[\textbf{$\star\star\star$}] Projects
\end{itemize}

\medskip
\textbf{Main references.}
\begin{itemize}
\item \textsc{Casella}, G. \& \textsc{Berger}, R.L. --- \emph{Statistical Inference}, 2nd ed., Cengage.
\item \textsc{Wasserman}, L. --- \emph{All of Statistics}, Springer.
\item \textsc{Saporta}, G. --- \emph{Probabilit\'es, Analyse des Donn\'ees et Statistique}, Technip.
\item \textsc{Wackerly}, Mendenhall \& Scheaffer --- \emph{Mathematical Statistics with Applications}.
\end{itemize}

\mainmatter

% ============================================================
% CHAPTERS
% ============================================================
\input{chapters/ch01_descriptive}
\input{chapters/ch02_model_sampling}
\input{chapters/ch03_point_estimation}
\input{chapters/ch04_estimator_properties}
\input{chapters/ch05_confidence_intervals}
\input{chapters/ch06_hypothesis_testing}
\input{chapters/ch07_classical_tests}
\input{chapters/ch08_simple_regression}
\input{chapters/ch09_multiple_regression_anova}
\input{chapters/ch10_bayesian}
\input{chapters/ch11_nonparametric}

% ============================================================
\backmatter

\appendix
\chapter{Formula Sheet}
\label{app:formulas}

\section{Common Probability Distributions}

\begin{center}
\begin{tabular}{llcc}
\toprule
\textbf{Distribution} & \textbf{Parameters} & \textbf{Mean} & \textbf{Variance} \\
\midrule
$\mathrm{Bin}(n,p)$ & $n\in\N^*,\; p\in[0,1]$ & $np$ & $np(1-p)$ \\
$\mathrm{Pois}(\lambda)$ & $\lambda>0$ & $\lambda$ & $\lambda$ \\
$\mathrm{Unif}([a,b])$ & $a<b$ & $\frac{a+b}{2}$ & $\frac{(b-a)^2}{12}$ \\
$\mathrm{Exp}(\lambda)$ & $\lambda>0$ & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ \\
$\mathcal{N}(\mu,\sigma^2)$ & $\mu\in\R,\;\sigma>0$ & $\mu$ & $\sigma^2$ \\
$\chi^2(k)$ & $k\in\N^*$ & $k$ & $2k$ \\
$t(k)$ & $k\in\N^*$ & $0$ $(k>1)$ & $\frac{k}{k-2}$ $(k>2)$ \\
$F(d_1,d_2)$ & $d_1,d_2\in\N^*$ & $\frac{d_2}{d_2-2}$ & --- \\
$\Gamma(\alpha,\beta)$ & $\alpha,\beta>0$ & $\frac{\alpha}{\beta}$ & $\frac{\alpha}{\beta^2}$ \\
\bottomrule
\end{tabular}
\end{center}

\section{Estimation Formulas}
\begin{align}
\bar{x} &= \frac{1}{n}\sum_{i=1}^n x_i &
s^2 &= \frac{1}{n-1}\sum_{i=1}^n (x_i-\bar{x})^2 \\
\thetahat_{\text{MLE}} &= \argmax_\theta \sum_{i=1}^n \ln f(x_i;\theta) &
\thetahat_{\text{MoM}} &: \E_\theta[X^k] = \frac{1}{n}\sum_{i=1}^n X_i^k
\end{align}

\section{Confidence Intervals}
\begin{align}
\text{Mean (}\sigma\text{ known)} &: \bar{X} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}} \\
\text{Mean (}\sigma\text{ unknown)} &: \bar{X} \pm t_{\alpha/2,n-1}\frac{S}{\sqrt{n}} \\
\text{Proportion} &: \phat \pm z_{\alpha/2}\sqrt{\frac{\phat(1-\phat)}{n}} \\
\text{Variance} &: \left[\frac{(n-1)S^2}{\chi^2_{\alpha/2,n-1}},\;\frac{(n-1)S^2}{\chi^2_{1-\alpha/2,n-1}}\right]
\end{align}

\section{Test Statistics}
\begin{align}
Z &= \frac{\bar{X}-\mu_0}{\sigma/\sqrt{n}} \sim \mathcal{N}(0,1) \\
T &= \frac{\bar{X}-\mu_0}{S/\sqrt{n}} \sim t(n-1) \\
\chi^2 &= \frac{(n-1)S^2}{\sigma_0^2} \sim \chi^2(n-1) \\
F &= \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2} \sim F(n_1-1,n_2-1)
\end{align}

\chapter{Statistical Tables}
\label{app:tables}

Statistical tables (normal, $t$, $\chi^2$, $F$) are available as digital appendices. In practice, use Python:

\begin{minted}{python}
from scipy import stats
stats.norm.ppf(0.975)       # z_{0.025} = 1.96
stats.t.ppf(0.975, df=10)   # t_{0.025,10}
stats.chi2.ppf(0.95, df=5)  # chi2_{0.05,5}
stats.f.ppf(0.95, 3, 20)    # F_{0.05,3,20}
\end{minted}

% --- Bibliography ---
\bibliographystyle{plainnat}
\begin{thebibliography}{9}
\bibitem{casella} \textsc{Casella}, G. \& \textsc{Berger}, R.L. (2002). \emph{Statistical Inference}. 2nd edition, Cengage Learning.
\bibitem{wasserman} \textsc{Wasserman}, L. (2004). \emph{All of Statistics: A Concise Course in Statistical Inference}. Springer.
\bibitem{saporta} \textsc{Saporta}, G. (2011). \emph{Probabilit\'es, Analyse des Donn\'ees et Statistique}. Technip, 3rd edition.
\bibitem{wackerly} \textsc{Wackerly}, D., \textsc{Mendenhall}, W. \& \textsc{Scheaffer}, R. (2008). \emph{Mathematical Statistics with Applications}. 7th edition, Cengage.
\end{thebibliography}

\printindex

\end{document}
