% ============================================================
% DATA ANALYSIS WITH PYTHON FOR HEALTH SPECIALISTS
% Lecture Notes (30 hours) — English version
% Compilation: xelatex -shell-escape notes.tex (2 passes)
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\setlength{\headheight}{14.5pt}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,shapes,fit,backgrounds}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[python]{python3=true}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames,svgnames]{xcolor}
\definecolor{steelblue}{RGB}{70,130,180}
\definecolor{healthgreen}{RGB}{34,139,34}
\definecolor{warnorange}{RGB}{255,140,0}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Custom environments ---
\newtcolorbox{clinicalnote}{
  colback=healthgreen!5,colframe=healthgreen!70,
  title={\faStethoscope\ Clinical context},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{pythontip}{
  colback=steelblue!5,colframe=steelblue!70,
  title={\faPython\ Python tip},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{warningbox}{
  colback=warnorange!5,colframe=warnorange!70,
  title={\faExclamationTriangle\ Caution},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{exercise}{
  colback=gray!5,colframe=gray!50,
  title={Exercise},
  fonttitle=\bfseries,breakable
}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex

% --- Icons ---
\usepackage{fontawesome5}

% --- Theorem environments ---
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}{Example}[chapter]

% ============================================================
\title{
  \vspace{-2cm}
  {\Huge\bfseries Data Analysis with Python}\\[6pt]
  {\LARGE for Health Specialists}\\[12pt]
  {\large A 30-hour practical course}\\[24pt]
  {\normalsize Ya\'{e} Ulrich Gaba}\\[4pt]
  {\small AIRINA Labs}
}
\date{2026}
\author{}

% ============================================================
\begin{document}

% --- Title page ---
\frontmatter
\maketitle

% --- Table of contents ---
\tableofcontents
\newpage

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This course is designed for health professionals --- physicians, nurses, epidemiologists, public health researchers, and health policy analysts --- who need to analyze data but have little or no programming experience. You do not need to become a software developer. You need to become someone who can load a dataset, clean it, explore it, test a hypothesis, build a simple predictive model, and communicate the results clearly.

We use Python because it is free, widely adopted in health research, and has an ecosystem of libraries specifically designed for the kinds of analyses health professionals need: survival analysis, epidemiological measures, geospatial mapping, and clinical prediction.

Every chapter follows the same rhythm: a brief explanation of the concept, a worked example on real health data, and exercises you complete in Jupyter notebooks. All datasets used in this course are freely available from the WHO, CDC, UCI Machine Learning Repository, and other open sources.

\paragraph{What this course is not.} This is not a statistics course (though we use statistics). It is not a computer science course (though we write code). It is a \emph{practical data analysis course} for people whose primary expertise is health, not programming.

\paragraph{Prerequisites.} None beyond basic computer literacy. Familiarity with descriptive statistics (mean, median, standard deviation) is helpful but will be reviewed.

\paragraph{Software.} All code runs in Jupyter notebooks, either locally (Anaconda) or in the cloud (Google Colab --- free, no installation required).

\vfill
\noindent Ya\'{e} Ulrich Gaba\\
Cotonou, 2026

% ============================================================
\mainmatter

\include{chapters/ch01_python_essentials}
\include{chapters/ch02_health_data_pandas}
\include{chapters/ch03_data_cleaning}
\include{chapters/ch04_descriptive_epi}
\include{chapters/ch05_visualization}
\include{chapters/ch06_hypothesis_testing}
\include{chapters/ch07_regression}
\include{chapters/ch08_ml_clinical}
\include{chapters/ch09_geospatial_temporal}
\include{chapters/ch10_capstone}

% ============================================================
\backmatter

\chapter*{Appendix A: Python installation guide}
\addcontentsline{toc}{chapter}{Appendix A: Python installation guide}

\section*{Option 1: Google Colab (recommended for beginners)}
Go to \url{https://colab.research.google.com}. Sign in with a Google account. No installation needed. All libraries used in this course are pre-installed.

\section*{Option 2: Anaconda (local installation)}
Download Anaconda from \url{https://www.anaconda.com/download}. Install with default settings. Open Jupyter Notebook from Anaconda Navigator.

\section*{Required libraries}
\begin{minted}{bash}
pip install pandas matplotlib seaborn scipy scikit-learn lifelines geopandas
\end{minted}

\chapter*{Appendix B: Health data sources}
\addcontentsline{toc}{chapter}{Appendix B: Health data sources}

\begin{longtable}{p{4cm}p{8cm}p{3cm}}
\toprule
\textbf{Source} & \textbf{Description} & \textbf{URL} \\
\midrule
WHO GHO & Global health indicators (190+ countries, 1000+ indicators) & \url{who.int/data/gho} \\
Gapminder & Health, income, population by country and year & \url{gapminder.org/data} \\
Our World in Data & COVID-19, mortality, disease burden, vaccination & \url{ourworldindata.org} \\
CDC NHANES & US national health survey (labs, questionnaires, exams) & \url{cdc.gov/nchs/nhanes} \\
UCI ML Repository & Curated ML datasets (heart disease, diabetes, cancer) & \url{archive.ics.uci.edu} \\
World Bank & Health expenditure, life expectancy, maternal mortality & \url{data.worldbank.org} \\
Africa CDC & African disease surveillance and outbreak data & \url{africacdc.org} \\
JHU COVID-19 & Global COVID-19 cases, deaths, recoveries (time series) & \url{github.com/CSSEGISandData} \\
DHS Program & Demographic and Health Surveys (50+ African countries) & \url{dhsprogram.com} \\
\bottomrule
\end{longtable}

\bibliographystyle{plainnat}
\bibliography{references}

\printindex

\end{document}
