% ============================================================
% DEEP LEARNING — Lecture Notes (Master)
% English version — Compilation: xelatex -shell-escape (3 passes)
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\setlength{\headheight}{14.5pt}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,shapes,fit,backgrounds,chains,decorations.pathreplacing,matrix}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[python]{python3=true}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames,svgnames]{xcolor}
\definecolor{steelblue}{RGB}{70,130,180}
\definecolor{deepblue}{RGB}{0,51,102}
\definecolor{deepred}{RGB}{153,0,0}
\definecolor{deepgreen}{RGB}{0,102,51}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex

% ============================================================
% Theorem environments
% ============================================================
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}[definition]{Example}
\newtheorem{exercise}{Exercise}[chapter]
\newtheorem{proposition}[definition]{Proposition}

\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}

\theoremstyle{remark}
\newtheorem{remark}[definition]{Remark}

% ============================================================
% Boxes
% ============================================================
\newtcolorbox{keyformulas}[1][]{colback=blue!5,colframe=NavyBlue,
  title=\textbf{Key Formulas},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{warning}[1][]{colback=red!5,colframe=BrickRed,
  title=\textbf{Warning},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{bestpractice}[1][]{colback=green!5,colframe=ForestGreen,
  title=\textbf{Best Practice},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{codeblock}[1][]{colback=gray!5,colframe=gray!80,
  title=\textbf{Code},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{codeoutput}[1][]{colback=Apricot!5,colframe=Apricot!80!black,
  title=\textbf{Output},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{intuition}[1][]{colback=yellow!5,colframe=Goldenrod,
  title=\textbf{Intuition},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{architecturebox}[1][]{colback=Cyan!5,colframe=MidnightBlue,
  title=\textbf{Architecture},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{mathbox}[1][]{colback=blue!3,colframe=RoyalBlue,
  title=\textbf{Mathematical Derivation},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{sota}[1][]{colback=Lavender!30,colframe=DarkOrchid,
  title=\textbf{State of the Art},fonttitle=\bfseries,breakable,#1}

% ============================================================
% Commands
% ============================================================
\newcommand{\py}[1]{\mintinline{python}{#1}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\KL}{\mathrm{KL}}
\newcommand{\softmax}{\mathrm{softmax}}
\newcommand{\sigmoid}{\sigma}
\newcommand{\relu}{\mathrm{ReLU}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

% ============================================================
\title{\Huge\textbf{Deep Learning}\\[1em]
\Large Lecture Notes --- Master}
\author{Mathematics and Artificial Intelligence}
\date{Academic Year 2025--2026}

\begin{document}
\frontmatter

\begin{titlepage}
\begin{tikzpicture}[remember picture, overlay]
  \fill[white] (current page.south west) rectangle (current page.north east);
  \fill[blue!80!black] (current page.center) circle (5cm);
  \node[anchor=center, text=white, font=\Huge\bfseries, text width=8cm, align=center]
    at ([yshift=0.5cm]current page.center) {Deep Learning};
  \node[anchor=center, text=orange!80!yellow, font=\Large]
    at ([yshift=-1.5cm]current page.center) {Lecture Notes};
  \node[anchor=north, text=blue!70!black, font=\Large\itshape]
    at ([yshift=-6.5cm]current page.center) {Ya\'e Ulrich Gaba};
  \node[anchor=north, text=blue!50!black]
    at ([yshift=-8cm]current page.center) {M1--M2 --- 2025--2026};
  \draw[orange!70!yellow, line width=3pt] ([yshift=-9cm, xshift=-4cm]current page.center) -- ([yshift=-9cm, xshift=4cm]current page.center);
  \node[anchor=center, text width=10cm, align=center, text=blue!40!black, font=\itshape]
    at ([yshift=-11cm]current page.center) {``Deep learning allows computational models that are composed of multiple processing layers to learn representations of data with multiple levels of abstraction.'' --- Yann LeCun};
  \node[anchor=south, text=gray, font=\small] at ([yshift=1cm]current page.south) {\today};
\end{tikzpicture}
\end{titlepage}

\tableofcontents

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Deep learning has fundamentally transformed our approach to artificial intelligence. In just a few years, deep neural networks have evolved from academic curiosities to ubiquitous tools: image recognition, machine translation, text generation, drug discovery, protein structure prediction.

\medskip
Behind these spectacular advances lie deep mathematical foundations. Deep learning sits at the crossroads of linear algebra, analysis, probability theory, and optimization. Understanding these foundations is essential for designing new architectures, diagnosing training issues, and pushing the boundaries of the state of the art.

\medskip
These notes are designed for Master's students in mathematics, machine learning, and data science. They assume familiarity with linear algebra, probability theory, and the fundamentals of machine learning. Each chapter follows a rigorous progression:
\begin{enumerate}
\item Mathematical derivation of key results.
\item PyTorch implementation with complete training loops.
\item Hyperparameter analysis and ablation studies.
\item Connection to the current state of the art.
\item Graded exercises: mathematical ($\star$), implementation ($\star\star$), research project ($\star\star\star$).
\end{enumerate}

\medskip
\textbf{Main references.}
\begin{itemize}
\item \textsc{Goodfellow}, Bengio \& Courville --- \emph{Deep Learning}, MIT Press, 2016 (freely available online).
\item \textsc{Bishop} \& Bishop --- \emph{Deep Learning: Foundations and Concepts}, Springer, 2024.
\item \textsc{Zhang}, Lipton, Li \& Smola --- \emph{Dive into Deep Learning}, d2l.ai (interactive).
\item \textsc{LeCun}, Bengio \& Hinton --- ``Deep learning'', \emph{Nature}, 2015.
\end{itemize}

\mainmatter

\input{chapters/ch01_neural_networks}
\input{chapters/ch02_backpropagation}
\input{chapters/ch03_regularization}
\input{chapters/ch04_cnn}
\input{chapters/ch05_rnn_lstm}
\input{chapters/ch06_attention}
\input{chapters/ch07_transformers}
\input{chapters/ch08_gan}
\input{chapters/ch09_vae}
\input{chapters/ch10_diffusion}
\input{chapters/ch11_theory}
\input{chapters/ch12_applications}

% ============================================================
\backmatter
\appendix
\chapter{Architecture Glossary}
\label{app:glossary}

\begin{longtable}{lp{10cm}}
\toprule
\textbf{Architecture} & \textbf{Description} \\
\midrule
\endhead
MLP & Multilayer Perceptron --- fully connected network \\
CNN & Convolutional Neural Network --- spatial feature extraction \\
RNN & Recurrent Neural Network --- sequence processing \\
LSTM & Long Short-Term Memory --- RNN with gating mechanisms \\
GRU & Gated Recurrent Unit --- simplified LSTM variant \\
Transformer & Pure attention architecture --- full parallelization \\
GAN & Generative Adversarial Network --- generator vs discriminator \\
VAE & Variational Autoencoder --- generation via variational inference \\
DDPM & Denoising Diffusion Probabilistic Model --- iterative denoising \\
ResNet & Residual Network --- skip connections \\
U-Net & Encoder-decoder with skip connections \\
ViT & Vision Transformer --- image patches as tokens \\
BERT & Bidirectional pre-trained Transformer \\
GPT & Autoregressive generative Transformer \\
\bottomrule
\end{longtable}

% --- Bibliography ---
\begin{thebibliography}{99}
\bibitem{goodfellow} \textsc{Goodfellow}, I., Bengio, Y. \& Courville, A. (2016). \emph{Deep Learning}. MIT Press.
\bibitem{bishop2024} \textsc{Bishop}, C.M. \& Bishop, H. (2024). \emph{Deep Learning: Foundations and Concepts}. Springer.
\bibitem{d2l} \textsc{Zhang}, A., Lipton, Z.C., Li, M. \& Smola, A.J. (2023). \emph{Dive into Deep Learning}. Cambridge University Press.
\bibitem{lecun2015} \textsc{LeCun}, Y., Bengio, Y. \& Hinton, G. (2015). Deep learning. \emph{Nature}, 521(7553), 436--444.
\bibitem{vaswani} \textsc{Vaswani}, A. et al. (2017). Attention is all you need. \emph{NeurIPS}.
\bibitem{he2016} \textsc{He}, K. et al. (2016). Deep residual learning. \emph{CVPR}.
\bibitem{kingma2014} \textsc{Kingma}, D.P. \& Welling, M. (2014). Auto-encoding variational Bayes. \emph{ICLR}.
\bibitem{goodfellowgan} \textsc{Goodfellow}, I. et al. (2014). Generative adversarial nets. \emph{NeurIPS}.
\bibitem{ho2020} \textsc{Ho}, J., Jain, A. \& Abbeel, P. (2020). Denoising diffusion probabilistic models. \emph{NeurIPS}.
\bibitem{kingmaadam} \textsc{Kingma}, D.P. \& Ba, J. (2015). Adam: a method for stochastic optimization. \emph{ICLR}.
\bibitem{srivastava} \textsc{Srivastava}, N. et al. (2014). Dropout: a simple way to prevent neural networks from overfitting. \emph{JMLR}.
\bibitem{ioffe} \textsc{Ioffe}, S. \& Szegedy, C. (2015). Batch normalization. \emph{ICML}.
\bibitem{hochreiter} \textsc{Hochreiter}, S. \& Schmidhuber, J. (1997). Long short-term memory. \emph{Neural Computation}.
\bibitem{bahdanau} \textsc{Bahdanau}, D., Cho, K. \& Bengio, Y. (2015). Neural machine translation by jointly learning to align and translate. \emph{ICLR}.
\bibitem{bahdanau2015neural} \textsc{Bahdanau}, D., Cho, K. \& Bengio, Y. (2015). Neural machine translation by jointly learning to align and translate. \emph{ICLR}.
\bibitem{srivastava2014dropout} \textsc{Srivastava}, N. et al. (2014). Dropout: a simple way to prevent neural networks from overfitting. \emph{JMLR}, 15(1), 1929--1958.
\bibitem{ioffe2015batch} \textsc{Ioffe}, S. \& Szegedy, C. (2015). Batch normalization: accelerating deep network training. \emph{ICML}.
\bibitem{ba2016layer} \textsc{Ba}, J.L., Kiros, J.R. \& Hinton, G.E. (2016). Layer normalization. \emph{arXiv:1607.06450}.
\bibitem{zhang2018mixup} \textsc{Zhang}, H. et al. (2018). mixup: beyond empirical risk minimization. \emph{ICLR}.
\bibitem{yun2019cutmix} \textsc{Yun}, S. et al. (2019). CutMix: regularization strategy to train strong classifiers. \emph{ICCV}.
\bibitem{loshchilov2019decoupled} \textsc{Loshchilov}, I. \& Hutter, F. (2019). Decoupled weight decay regularization. \emph{ICLR}.
\bibitem{huang2016deep} \textsc{Huang}, G. et al. (2017). Densely connected convolutional networks. \emph{CVPR}.
\bibitem{szegedy2016rethinking} \textsc{Szegedy}, C. et al. (2016). Rethinking the Inception architecture. \emph{CVPR}.
\bibitem{wu2021r} \textsc{Wu}, H. et al. (2021). CvT: introducing convolutions to Vision Transformers. \emph{ICCV}.
\bibitem{krizhevsky2012imagenet} \textsc{Krizhevsky}, A., Sutskever, I. \& Hinton, G.E. (2012). ImageNet classification with deep convolutional neural networks. \emph{NeurIPS}.
\bibitem{simonyan2015very} \textsc{Simonyan}, K. \& Zisserman, A. (2015). Very deep convolutional networks for large-scale image recognition. \emph{ICLR}.
\bibitem{tan2019efficientnet} \textsc{Tan}, M. \& Le, Q.V. (2019). EfficientNet: rethinking model scaling for convolutional neural networks. \emph{ICML}.
\bibitem{liu2022convnet} \textsc{Liu}, Z. et al. (2022). A ConvNet for the 2020s. \emph{CVPR}.
\bibitem{dai2017deformable} \textsc{Dai}, J. et al. (2017). Deformable convolutional networks. \emph{ICCV}.
\bibitem{hochreiter1997long} \textsc{Hochreiter}, S. \& Schmidhuber, J. (1997). Long short-term memory. \emph{Neural Computation}, 9(8), 1735--1780.
\bibitem{cho2014learning} \textsc{Cho}, K. et al. (2014). Learning phrase representations using RNN encoder-decoder. \emph{EMNLP}.
\bibitem{gu2024mamba} \textsc{Gu}, A. \& Dao, T. (2024). Mamba: linear-time sequence modeling with selective state spaces. \emph{ICLR}.
\bibitem{beck2024xlstm} \textsc{Beck}, M. et al. (2024). xLSTM: extended Long Short-Term Memory. \emph{arXiv:2405.04517}.
\bibitem{sutskever2014sequence} \textsc{Sutskever}, I., Vinyals, O. \& Le, Q.V. (2014). Sequence to sequence learning with neural networks. \emph{NeurIPS}.
\bibitem{luong2015effective} \textsc{Luong}, M.-T., Pham, H. \& Manning, C.D. (2015). Effective approaches to attention-based neural machine translation. \emph{EMNLP}.
\bibitem{vaswani2017attention} \textsc{Vaswani}, A. et al. (2017). Attention is all you need. \emph{NeurIPS}.
\bibitem{dao2022flashattention} \textsc{Dao}, T. et al. (2022). FlashAttention: fast and memory-efficient exact attention. \emph{NeurIPS}.
\bibitem{child2019generating} \textsc{Child}, R. et al. (2019). Generating long sequences with sparse transformers. \emph{arXiv:1904.10509}.
\bibitem{shazeer2019fast} \textsc{Shazeer}, N. (2019). Fast transformer decoding: one write-head is all you need. \emph{arXiv:1911.02150}.
\bibitem{ye2024differential} \textsc{Ye}, W. et al. (2024). Differential Transformer. \emph{arXiv:2410.05258}.
\bibitem{choromanski2021rethinking} \textsc{Choromanski}, K. et al. (2021). Rethinking attention with Performers. \emph{ICLR}.
\end{thebibliography}

\printindex
\end{document}
