% ============================================================
\chapter{ETL and Data Warehousing}
\label{ch:etl}
% ============================================================

\begin{intuition}[title=\textbf{Central idea}]
A data warehouse centralizes enterprise data in a schema optimized for
analytical queries. ETL (Extract, Transform, Load) pipelines feed this
warehouse by cleaning and transforming data from heterogeneous sources.
\end{intuition}

% ------------------------------------------------------------
\section{Fundamental concepts}
% ------------------------------------------------------------

\begin{definition}[Data warehouse]
\index{data warehouse}
A \textbf{data warehouse} is a database that is \emph{subject-oriented},
\emph{integrated}, \emph{non-volatile}, and \emph{time-variant}, designed
for decision support (Inmon, 1992).
\end{definition}

\begin{remark}
A data warehouse contrasts with \textbf{OLTP} (\emph{Online Transaction
Processing}) databases that serve daily operations. Analytical workloads
fall under \textbf{OLAP} (\emph{Online Analytical Processing}).
\end{remark}

\begin{center}
\begin{tabular}{lcc}
\hline
\textbf{Characteristic} & \textbf{OLTP} & \textbf{OLAP} \\
\hline
Operations & Short reads/writes & Long analytical queries \\
Users & Many, operational & Few, analysts \\
Data & Current, detailed & Historical, aggregated \\
Schema & Normalized (3NF) & Denormalized (star, snowflake) \\
\hline
\end{tabular}
\end{center}

% ------------------------------------------------------------
\section{Dimensional modeling}
% ------------------------------------------------------------

\begin{definition}[Star schema]
\index{star schema}
A \textbf{star schema} organizes data around a central \textbf{fact table}
(quantitative measures) surrounded by \textbf{dimension tables}
(descriptive attributes). Dimensions are linked to the fact table via
foreign keys.
\end{definition}

\begin{example}
Data warehouse for retail sales:
\begin{minted}{python}
-- Fact table
CREATE TABLE fact_sales (
    sale_id      SERIAL PRIMARY KEY,
    product_id   INT REFERENCES dim_product(id),
    customer_id  INT REFERENCES dim_customer(id),
    date_id      INT REFERENCES dim_date(id),
    store_id     INT REFERENCES dim_store(id),
    quantity     INT,
    amount       DECIMAL(10,2),
    cost         DECIMAL(10,2)
);

-- Dimension table
CREATE TABLE dim_date (
    id        INT PRIMARY KEY,
    full_date DATE,
    month     INT,
    quarter   INT,
    year      INT,
    day_of_week VARCHAR(10)
);
\end{minted}
\end{example}

\begin{definition}[Snowflake schema]
\index{snowflake schema}
The \textbf{snowflake schema} is a variant where dimension tables are
normalized: each hierarchy is extracted into a separate table
(e.g., \texttt{dim\_city} $\to$ \texttt{dim\_region} $\to$
\texttt{dim\_country}).
\end{definition}

\begin{attention}
The star schema is generally preferred over the snowflake because:
\begin{itemize}
  \item joins are simpler (single step),
  \item performance is better (fewer joins),
  \item redundancy is acceptable in OLAP (disk space is cheap).
\end{itemize}
\end{attention}

% ------------------------------------------------------------
\section{The ETL process}
% ------------------------------------------------------------

\begin{definition}[ETL]
\index{ETL}
The \textbf{ETL} (\emph{Extract, Transform, Load}) process consists of:
\begin{enumerate}
  \item \textbf{Extract}: retrieve data from sources (OLTP databases,
        CSV files, APIs, logs).
  \item \textbf{Transform}: clean, normalize, deduplicate, aggregate,
        compute derived metrics.
  \item \textbf{Load}: insert the transformed data into the warehouse.
\end{enumerate}
\end{definition}

\begin{example}
ETL pipeline in Python with pandas:
\begin{minted}{python}
import pandas as pd
from sqlalchemy import create_engine

# 1. Extract
df_sales = pd.read_csv("sales_2025.csv")
df_customers = pd.read_sql("SELECT * FROM customers", engine_oltp)

# 2. Transform
df_sales["date"] = pd.to_datetime(df_sales["date"])
df_sales = df_sales.dropna(subset=["amount"])
df_sales["amount"] = df_sales["amount"].clip(lower=0)
df = df_sales.merge(df_customers, on="customer_id", how="left")
df["quarter"] = df["date"].dt.quarter

# 3. Load
engine_dw = create_engine("postgresql://user:pass@host/warehouse")
df.to_sql("fact_sales", engine_dw, if_exists="append", index=False)
\end{minted}
\end{example}

\begin{definition}[ELT]
\index{ELT}
The \textbf{ELT} (\emph{Extract, Load, Transform}) variant loads raw data
into the warehouse first, then performs transformations directly in SQL.
This approach is popular with cloud warehouses (BigQuery, Snowflake,
Redshift) thanks to their compute power.
\end{definition}

% ------------------------------------------------------------
\section{Slowly Changing Dimensions (SCD)}
% ------------------------------------------------------------

\begin{definition}[Slowly Changing Dimensions]
\index{SCD}
\textbf{SCDs} manage modifications to dimension attributes:
\begin{itemize}
  \item \textbf{Type 1}: overwrite the old value (no history).
  \item \textbf{Type 2}: create a new row with validity dates
        (\texttt{start\_date}, \texttt{end\_date}, \texttt{is\_current}).
  \item \textbf{Type 3}: add a column for the old value (limited history).
\end{itemize}
\end{definition}

\begin{example}
SCD Type 2 --- a customer moves:
\begin{minted}{python}
-- Before: current row
-- id=1, name='Smith', city='New York', start_date='2020-01-01',
--        end_date='9999-12-31', is_current=TRUE

-- After relocation:
UPDATE dim_customer SET end_date = '2025-06-15', is_current = FALSE
WHERE id = 1 AND is_current = TRUE;

INSERT INTO dim_customer (natural_id, name, city, start_date, end_date, is_current)
VALUES (1, 'Smith', 'Boston', '2025-06-15', '9999-12-31', TRUE);
\end{minted}
\end{example}

% ------------------------------------------------------------
\section{OLAP operations}
% ------------------------------------------------------------

\begin{definition}[OLAP operations]
\index{OLAP}
The classic OLAP operations on a multidimensional cube are:
\begin{itemize}
  \item \textbf{Roll-up} (aggregation): move up in the hierarchy
        (day $\to$ month $\to$ year).
  \item \textbf{Drill-down} (disaggregation): move down in the hierarchy.
  \item \textbf{Slice}: fix one dimension (e.g., \texttt{year = 2025}).
  \item \textbf{Dice}: select a multi-dimensional sub-cube.
  \item \textbf{Pivot}: rotate the axes of the cube.
\end{itemize}
\end{definition}

\begin{minted}{python}
-- Roll-up: sales by month -> sales by quarter
SELECT d.quarter, d.year, SUM(f.amount) AS total
FROM fact_sales f
JOIN dim_date d ON f.date_id = d.id
GROUP BY d.quarter, d.year
ORDER BY d.year, d.quarter;

-- Slice + Dice with CUBE
SELECT d.year, p.category, s.region,
       SUM(f.amount) AS total
FROM fact_sales f
JOIN dim_date d ON f.date_id = d.id
JOIN dim_product p ON f.product_id = p.id
JOIN dim_store s ON f.store_id = s.id
WHERE d.year IN (2024, 2025)
GROUP BY CUBE (d.year, p.category, s.region);
\end{minted}

% ------------------------------------------------------------
\section{Materialized views}
% ------------------------------------------------------------

\begin{definition}[Materialized view]
\index{materialized view}
A \textbf{materialized view} is a view whose result is physically stored
on disk. It accelerates repetitive analytical queries at the cost of
additional storage and a refresh overhead.
\end{definition}

\begin{minted}{python}
-- Creating a materialized view
CREATE MATERIALIZED VIEW mv_monthly_sales AS
SELECT d.year, d.month, p.category,
       SUM(f.amount) AS total, COUNT(*) AS num_sales
FROM fact_sales f
JOIN dim_date d ON f.date_id = d.id
JOIN dim_product p ON f.product_id = p.id
GROUP BY d.year, d.month, p.category;

-- Refreshing
REFRESH MATERIALIZED VIEW CONCURRENTLY mv_monthly_sales;
\end{minted}

% ------------------------------------------------------------
\section{Key formulas}
% ------------------------------------------------------------

\begin{keyformulas}
\begin{itemize}
  \item \textbf{ETL}: Extract $\to$ Transform $\to$ Load (vs.\ ELT: Extract $\to$ Load $\to$ Transform)
  \item \textbf{Star schema}: 1 fact table + $k$ dimension tables, simple joins
  \item \textbf{OLAP}: Roll-up, Drill-down, Slice, Dice, Pivot
  \item \textbf{SCD Type 2}: historization via \texttt{start\_date}/\texttt{end\_date}
  \item \textbf{Materialized view}: space vs.\ query time trade-off
\end{itemize}
\end{keyformulas}

% ------------------------------------------------------------
\section{Exercises}
% ------------------------------------------------------------

\begin{exercise}
Design a star schema for an airline reservation system. Identify the fact
table, measures, and at least four dimensions.
\end{exercise}

\begin{exercise}
Transform the star schema from the previous exercise into a snowflake
schema. Discuss the advantages and disadvantages.
\end{exercise}

\begin{exercise}
Write an ETL pipeline in Python that: (a) extracts data from a JSON file,
(b) cleans missing values and normalizes dates, (c) loads the result into
a PostgreSQL table.
\end{exercise}

\begin{exercise}
Implement a Type 2 slowly changing dimension for the \texttt{dim\_product}
table. Simulate a price change and verify that history is preserved.
\end{exercise}

\begin{exercise}
Write the SQL queries corresponding to the following OLAP operations on the
sales schema: (a) roll-up from day to quarter, (b) slice on the year 2025,
(c) dice on electronics products sold in California.
\end{exercise}
