% ============================================================
% NUMERICAL ANALYSIS — Lecture Notes (L2/L3)
% English version — Compilation: xelatex -shell-escape
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,decorations.pathreplacing,shapes}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[python]{python3=true}

% --- Algorithms ---
\usepackage[ruled,vlined]{algorithm2e}
\SetKwInput{KwInput}{Input}
\SetKwInput{KwOutput}{Output}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames]{xcolor}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex

% ============================================================
% Theorem environments
% ============================================================
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}[definition]{Example}
\newtheorem{exercise}{Exercise}[chapter]

\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}

\theoremstyle{remark}
\newtheorem{remark}[definition]{Remark}

% ============================================================
% Boxes
% ============================================================
\newtcolorbox{keyformulas}[1][]{colback=blue!5,colframe=NavyBlue,
  title=\textbf{Key Formulas},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{warning}[1][]{colback=red!5,colframe=BrickRed,
  title=\textbf{Warning},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{method}[1][]{colback=green!5,colframe=ForestGreen,
  title=\textbf{Method},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{codeblock}[1][]{colback=gray!5,colframe=gray!80,
  title=\textbf{Implementation},fonttitle=\bfseries,breakable,#1}

% ============================================================
% Commands
% ============================================================
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\eps}{\varepsilon}
\newcommand{\fl}{\mathrm{fl}}
\newcommand{\cond}{\mathrm{cond}}
\newcommand{\BigO}{\mathcal{O}}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\spn}{span}
\DeclareMathOperator{\rang}{rank}

% ============================================================
\begin{document}
\frontmatter
\begin{titlepage}
\begin{tikzpicture}[remember picture, overlay]
  \fill[blue!10!white] (current page.south west) rectangle (current page.north east);
  \fill[blue!80!black, rotate around={-15:(current page.center)}]
    ([xshift=-4cm,yshift=-1cm]current page.center) rectangle ([xshift=4cm,yshift=1.5cm]current page.center);
  \node[anchor=center, text=white, font=\Huge\bfseries, rotate=-15]
    at ([yshift=0.3cm]current page.center) {Numerical Analysis};
  \node[anchor=center, text=orange!80!yellow, font=\Large, rotate=-15]
    at ([yshift=-1cm]current page.center) {Lecture Notes};
  \node[anchor=north, text=blue!70!black, font=\Large\itshape]
    at ([yshift=-4cm]current page.center) {Ya\'e Ulrich Gaba};
  \node[anchor=north, text=blue!50!black, font=\large]
    at ([yshift=-5.5cm]current page.center) {Licence L3 --- 2025--2026};
  \fill[orange!70!yellow] ([yshift=3cm]current page.south west) rectangle ([yshift=3.3cm]current page.south east);
  \node[anchor=south, text=gray, font=\small] at ([yshift=1cm]current page.south) {\today};
\end{tikzpicture}
\end{titlepage}
\tableofcontents
\listofalgorithms
\addcontentsline{toc}{chapter}{List of Algorithms}

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Numerical analysis is the art of solving mathematical problems \emph{approximately but with controlled error}. Unlike theoretical analysis, which establishes existence and uniqueness of solutions, numerical analysis is concerned with \emph{actually computing} them and \emph{quantifying the error incurred}.

\medskip
This course covers the fundamentals: floating-point arithmetic, solving linear systems, interpolation, numerical integration, ordinary differential equations, and eigenvalue problems. Each method is accompanied by its error analysis, pseudocode, and implementations in Python and Julia.

\medskip
\textbf{Prerequisites.} Real Analysis I \& II, linear algebra, basic programming.

\medskip
\textbf{References.}
\begin{itemize}
\item \textsc{Quarteroni}, Sacco, Saleri --- \emph{Numerical Mathematics}, Springer.
\item \textsc{S\"uli}, Mayers --- \emph{An Introduction to Numerical Analysis}, Cambridge.
\item \textsc{Trefethen}, Bau --- \emph{Numerical Linear Algebra}, SIAM.
\item \textsc{Higham} --- \emph{Accuracy and Stability of Numerical Algorithms}, SIAM.
\item \textsc{Demailly} --- \emph{Analyse Num\'erique et \'Equations Diff\'erentielles}, EDP Sciences.
\end{itemize}

\mainmatter

\input{chapters/ch01_floating_point}
\input{chapters/ch02_root_finding}
\input{chapters/ch03_direct_methods}
\input{chapters/ch04_iterative_methods}
\input{chapters/ch05_interpolation}
\input{chapters/ch06_approximation}
\input{chapters/ch07_differentiation_integration}
\input{chapters/ch08_gauss_quadrature}
\input{chapters/ch09_ode}
\input{chapters/ch10_eigenvalues}

% ============================================================
\backmatter
\appendix
\chapter{Formula Sheet}
\label{app:formulas}

\section{Vector and Matrix Norms}
\begin{align}
\|x\|_1 &= \sum|x_i|, \quad \|x\|_2=\sqrt{\sum x_i^2}, \quad \|x\|_\infty=\max|x_i| \\
\|A\|_1 &= \max_j\sum_i|a_{ij}|, \quad \|A\|_\infty=\max_i\sum_j|a_{ij}|, \quad \|A\|_2=\sigma_{\max}(A)
\end{align}

\section{Errors}
\begin{align}
\text{Absolute error} &: |x-\tilde{x}| \\
\text{Relative error} &: \frac{|x-\tilde{x}|}{|x|} \\
\text{Condition number} &: \cond(A)=\|A\|\cdot\|A^{-1}\|
\end{align}

\section{Quadrature Rules}
\begin{align}
\text{Trapezoidal} &: \frac{h}{2}[f(a)+2\sum f(x_i)+f(b)], \quad E=\BigO(h^2) \\
\text{Simpson} &: \frac{h}{3}[f(a)+4\sum f_{\text{odd}}+2\sum f_{\text{even}}+f(b)], \quad E=\BigO(h^4)
\end{align}

\section{ODE Methods}
\begin{align}
\text{Explicit Euler} &: y_{n+1}=y_n+hf(t_n,y_n), \quad \BigO(h) \\
\text{RK4} &: y_{n+1}=y_n+\tfrac{h}{6}(k_1+2k_2+2k_3+k_4), \quad \BigO(h^4)
\end{align}

% --- Bibliography ---
\begin{thebibliography}{9}
\bibitem{quarteroni} \textsc{Quarteroni}, A., Sacco, R. \& Saleri, F. (2007). \emph{Numerical Mathematics}. 2nd ed., Springer.
\bibitem{suli} \textsc{S\"uli}, E. \& Mayers, D. (2003). \emph{An Introduction to Numerical Analysis}. Cambridge University Press.
\bibitem{trefethen} \textsc{Trefethen}, L.N. \& Bau, D. (1997). \emph{Numerical Linear Algebra}. SIAM.
\bibitem{higham} \textsc{Higham}, N.J. (2002). \emph{Accuracy and Stability of Numerical Algorithms}. 2nd ed., SIAM.
\bibitem{demailly} \textsc{Demailly}, J.-P. (2006). \emph{Analyse Num\'erique et \'Equations Diff\'erentielles}. EDP Sciences.
\end{thebibliography}

\printindex
\end{document}
