% ============================================================
% INTRODUCTION TO DATA SCIENCE — Lecture Notes (Undergraduate)
% English version — Compilation: xelatex -shell-escape (3 passes)
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\setlength{\headheight}{14.5pt}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,shapes,fit,backgrounds}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[python]{python3=true}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames,svgnames]{xcolor}
\definecolor{steelblue}{RGB}{70,130,180}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex

% ============================================================
% Theorem environments
% ============================================================
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}[definition]{Example}
\newtheorem{exercise}{Exercise}[chapter]

\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{proposition}[definition]{Proposition}

\theoremstyle{remark}
\newtheorem{remark}[definition]{Remark}

% ============================================================
% Boxes
% ============================================================
\newtcolorbox{keyformulas}[1][]{colback=blue!5,colframe=NavyBlue,
  title=\textbf{Key Functions},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{warning}[1][]{colback=red!5,colframe=BrickRed,
  title=\textbf{Warning},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{bestpractice}[1][]{colback=green!5,colframe=ForestGreen,
  title=\textbf{Best Practice},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{codeblock}[1][]{colback=gray!5,colframe=gray!80,
  title=\textbf{Code},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{codeoutput}[1][]{colback=Apricot!5,colframe=Apricot!80!black,
  title=\textbf{Output},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{intuition}[1][]{colback=yellow!5,colframe=Goldenrod,
  title=\textbf{Intuition},fonttitle=\bfseries,breakable,#1}

% ============================================================
% Commands
% ============================================================
\newcommand{\py}[1]{\mintinline{python}{#1}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\operatorname{Var}}

% ============================================================
\begin{document}
\frontmatter
\begin{titlepage}
\begin{tikzpicture}[remember picture, overlay]
  \fill[white] (current page.south west) rectangle (current page.north east);
  \fill[blue!80!black] (current page.center) circle (5cm);
  \node[anchor=center, text=white, font=\Huge\bfseries, text width=8cm, align=center]
    at ([yshift=0.5cm]current page.center) {Introduction to Data Science};
  \node[anchor=center, text=orange!80!yellow, font=\Large]
    at ([yshift=-1.5cm]current page.center) {Lecture Notes};
  \node[anchor=north, text=blue!70!black, font=\Large\itshape]
    at ([yshift=-6.5cm]current page.center) {Ya\'e Ulrich Gaba};
  \node[anchor=north, text=blue!50!black]
    at ([yshift=-8cm]current page.center) {Licence L3 --- 2025--2026};
  \draw[orange!70!yellow, line width=3pt] ([yshift=-9cm, xshift=-4cm]current page.center) -- ([yshift=-9cm, xshift=4cm]current page.center);
  \node[anchor=center, text width=10cm, align=center, text=blue!40!black, font=\itshape]
    at ([yshift=-11cm]current page.center) {``All models are wrong, but some are useful.''\\[4pt]--- George E. P. Box};
  \node[anchor=south, text=gray, font=\small] at ([yshift=1cm]current page.south) {\today};
\end{tikzpicture}
\end{titlepage}
\tableofcontents

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\textbf{Data science} is an interdisciplinary field that combines mathematics, statistics, and computer science to extract knowledge from data. It differs from classical statistics in scale (data volume), tools (programming, machine learning), and practical orientation (decisions, products, predictions).

\medskip
These notes are designed for undergraduate students with basic Python and statistics knowledge. Each chapter starts from a \textbf{real dataset}, introduces concepts, and implements them in Python. The goal is to make the student autonomous in conducting a complete data analysis, from import to communication of results.

\medskip
\textbf{Prerequisites.} Basic Python programming (variables, loops, functions), introductory descriptive statistics and probability.

\medskip
\textbf{Datasets used.} Iris, Titanic, Tips (Seaborn), California Housing, 20 Newsgroups, AirPassengers --- all freely available.

\medskip
\textbf{References.}
\begin{itemize}
\item \textsc{VanderPlas} --- \emph{Python Data Science Handbook}, O'Reilly (freely available online).
\item \textsc{McKinney} --- \emph{Python for Data Analysis}, O'Reilly.
\item \textsc{G\'eron} --- \emph{Hands-On Machine Learning with Scikit-Learn, Keras, and TensorFlow}, O'Reilly.
\item \textsc{James}, Witten, Hastie, Tibshirani --- \emph{An Introduction to Statistical Learning (ISLR)}, Springer (free).
\end{itemize}

\mainmatter

\input{chapters/ch01_pipeline}
\input{chapters/ch02_pandas}
\input{chapters/ch03_visualization}
\input{chapters/ch04_eda}
\input{chapters/ch05_cleaning}
\input{chapters/ch06_statistics}
\input{chapters/ch07_ml_intro}
\input{chapters/ch08_models}
\input{chapters/ch09_evaluation}
\input{chapters/ch10_text_temporal}
\input{chapters/ch11_projects}

% ============================================================
\backmatter
\appendix
\chapter{Python Quick Reference}
\label{app:python}

\section{Pandas}
\begin{tabular}{ll}
\toprule
\textbf{Function} & \textbf{Description} \\
\midrule
\py{pd.read_csv()} & Load a CSV file \\
\py{df.head()} & First rows \\
\py{df.describe()} & Descriptive statistics \\
\py{df.groupby()} & Aggregate by groups \\
\py{df.merge()} & Join two DataFrames \\
\py{df.pivot_table()} & Pivot table \\
\py{df.dropna()} / \py{df.fillna()} & Handle missing values \\
\bottomrule
\end{tabular}

\section{Visualization}
\begin{tabular}{ll}
\toprule
\textbf{Function} & \textbf{Description} \\
\midrule
\py{plt.plot()} & Line plot \\
\py{plt.scatter()} & Scatter plot \\
\py{plt.hist()} & Histogram \\
\py{plt.bar()} & Bar chart \\
\py{sns.heatmap()} & Heatmap \\
\py{sns.pairplot()} & Pair plot \\
\bottomrule
\end{tabular}

\section{Scikit-learn}
\begin{tabular}{ll}
\toprule
\textbf{Class / Function} & \textbf{Description} \\
\midrule
\py{train_test_split()} & Train/test split \\
\py{LinearRegression()} & Linear regression \\
\py{LogisticRegression()} & Logistic regression \\
\py{KNeighborsClassifier()} & $k$-nearest neighbors \\
\py{DecisionTreeClassifier()} & Decision tree \\
\py{KMeans()} & $k$-means clustering \\
\py{cross_val_score()} & Cross-validation \\
\bottomrule
\end{tabular}

% --- Bibliography ---
\begin{thebibliography}{9}
\bibitem{vanderplas} \textsc{VanderPlas}, J. (2016). \emph{Python Data Science Handbook}. O'Reilly.
\bibitem{mckinney} \textsc{McKinney}, W. (2022). \emph{Python for Data Analysis}. 3rd ed., O'Reilly.
\bibitem{geron} \textsc{G\'eron}, A. (2022). \emph{Hands-On Machine Learning}. 3rd ed., O'Reilly.
\bibitem{islr} \textsc{James}, G. et al. (2021). \emph{An Introduction to Statistical Learning}. 2nd ed., Springer.
\end{thebibliography}

\printindex
\end{document}
