---
title: "Day 1: R Fundamentals"
subtitle: "R for Statistical Analysis Workshop"
author: "Dr. Yae Ulrich Gaba"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Welcome to Day 1 of the **R for Statistical Analysis** workshop. Today we cover
the building blocks of R programming: data types, data structures, control flow,
functions, and pipes. By the end of this session you will be comfortable working
interactively in R and writing small scripts.

---

# R as a Calculator

R can be used as a powerful calculator right out of the box.

```{r arithmetic}
# Basic arithmetic
2 + 3
10 - 4
6 * 7
45 / 9
2^10          # exponentiation
17 %% 5      # modulo (remainder)
17 %/% 5     # integer division
```

```{r math-functions}
# Useful math functions
sqrt(144)
abs(-7.5)
log(100)        # natural log
log10(100)      # base-10 log
exp(1)          # e^1
round(3.14159, 2)
ceiling(4.3)
floor(4.9)
```

---

# Variables and Assignment

In R we assign values to variables with `<-` (preferred) or `=`.

```{r variables}
x <- 42
y = 3.14
name <- "Gaba"
is_workshop <- TRUE

x
y
name
is_workshop
```

```{r variable-types}
# Check types
class(x)
class(name)
class(is_workshop)
typeof(x)
typeof(name)
```

---

# Vectors

Vectors are the most fundamental data structure in R. Every vector has a single
type.

## Numeric Vectors

```{r numeric-vectors}
# Creating numeric vectors
ages <- c(23, 31, 45, 28, 37)
ages

# Sequences
1:10
seq(0, 1, by = 0.2)
seq(1, 100, length.out = 5)

# Repetition
rep(0, 5)
rep(c(1, 2, 3), times = 3)
rep(c(1, 2, 3), each = 3)
```

## Character Vectors

```{r character-vectors}
countries <- c("Benin", "Togo", "Ghana", "Nigeria", "Senegal")
countries

# Useful string operations
nchar(countries)
toupper(countries)
paste("Country:", countries)
paste0("ID_", 1:5)
```

## Logical Vectors

```{r logical-vectors}
passed <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
passed

# Logical operations
!passed
passed & c(TRUE, TRUE, FALSE, FALSE, TRUE)
passed | c(FALSE, FALSE, FALSE, TRUE, TRUE)

# Logical vectors from comparisons
ages > 30
ages >= 28 & ages <= 40
```

## Factors

Factors represent categorical data with a fixed set of levels.

```{r factors}
education <- c("BSc", "MSc", "PhD", "BSc", "PhD", "MSc", "BSc")
edu_factor <- factor(education)
edu_factor
levels(edu_factor)
table(edu_factor)

# Ordered factor
edu_ordered <- factor(education,
                      levels = c("BSc", "MSc", "PhD"),
                      ordered = TRUE)
edu_ordered
edu_ordered[1] < edu_ordered[3]
```

## Vector Indexing

```{r indexing}
vals <- c(10, 20, 30, 40, 50, 60, 70, 80)

# Positive indexing
vals[1]
vals[c(1, 3, 5)]
vals[2:4]

# Negative indexing (exclude)
vals[-1]
vals[-c(1, 2)]

# Logical indexing
vals[vals > 40]
vals[vals >= 30 & vals <= 60]

# Named indexing
scores <- c(math = 90, physics = 85, chemistry = 92)
scores["physics"]
scores[c("math", "chemistry")]
```

## Vectorised Operations

```{r vectorised}
a <- c(1, 2, 3, 4, 5)
b <- c(10, 20, 30, 40, 50)

a + b
a * b
a^2
sum(a)
mean(a)
sd(a)
cumsum(a)
```

---

# Matrices

Matrices are 2-dimensional arrays of a single type.

```{r matrices}
# Create a matrix
m <- matrix(1:12, nrow = 3, ncol = 4)
m

# With row and column names
m2 <- matrix(c(85, 90, 78, 92, 88, 76, 95, 80, 82),
             nrow = 3, byrow = TRUE,
             dimnames = list(c("Alice", "Bob", "Carol"),
                             c("Math", "Physics", "Chemistry")))
m2

# Dimensions
dim(m2)
nrow(m2)
ncol(m2)
```

```{r matrix-indexing}
# Indexing
m2[1, 2]             # row 1, col 2
m2["Bob", ]          # entire row
m2[, "Chemistry"]    # entire column
m2[c(1, 3), c(1, 3)] # sub-matrix

# Matrix operations
t(m2)                # transpose
colMeans(m2)
rowSums(m2)
```

---

# Lists

Lists can hold elements of different types and lengths.

```{r lists}
student <- list(
  name = "Amina",
  age = 24,
  scores = c(88, 92, 79),
  graduated = FALSE
)
student

# Accessing elements
student$name
student[["age"]]
student[[3]]        # third element (scores)

# Length
length(student)
```

```{r nested-lists}
# Nested list
department <- list(
  name = "Statistics",
  head = "Dr. Gaba",
  students = list(
    list(name = "Kofi", gpa = 3.8),
    list(name = "Fatou", gpa = 3.5),
    list(name = "Emeka", gpa = 3.9)
  )
)

department$students[[2]]$name
department$students[[3]]$gpa
```

---

# Data Frames

Data frames are the workhorse for tabular data in R.

```{r data-frames}
df <- data.frame(
  name = c("Amina", "Kofi", "Fatou", "Emeka", "Ngozi"),
  age = c(24, 27, 22, 30, 25),
  country = c("Nigeria", "Ghana", "Senegal", "Nigeria", "Nigeria"),
  score = c(88, 76, 92, 85, 90),
  stringsAsFactors = FALSE
)
df

str(df)
summary(df)
```

```{r df-operations}
# Accessing columns
df$name
df[, "score"]
df[["age"]]

# Filtering rows
df[df$score > 85, ]
df[df$country == "Nigeria", ]

# Adding a new column
df$pass <- df$score >= 80
df

# Dimensions
nrow(df)
ncol(df)
names(df)
```

## Tibbles

Tibbles are a modern take on data frames (from the `tibble` package).

```{r tibbles}
library(tibble)

tb <- tibble(
  id = 1:5,
  value = rnorm(5),
  group = sample(c("A", "B"), 5, replace = TRUE)
)
tb

# Tibbles print nicely and never convert strings to factors
class(tb)
```

---

# Control Flow

## if / else

```{r if-else}
temperature <- 35

if (temperature > 30) {
  message("It is hot today.")
} else if (temperature > 20) {
  message("Pleasant weather.")
} else {
  message("It is cool.")
}
```

```{r ifelse-vectorised}
# Vectorised ifelse
scores <- c(45, 72, 88, 55, 91)
result <- ifelse(scores >= 60, "Pass", "Fail")
result
```

## for Loops

```{r for-loop}
# Simple loop
for (i in 1:5) {
  cat("Iteration", i, "- square:", i^2, "\n")
}
```

```{r for-loop-accumulate}
# Accumulating results
countries <- c("Benin", "Togo", "Ghana")
greetings <- character(length(countries))

for (i in seq_along(countries)) {
  greetings[i] <- paste("Welcome to", countries[i])
}
greetings
```

## while Loops

```{r while-loop}
count <- 1
total <- 0

while (total < 50) {
  total <- total + count
  count <- count + 1
}
cat("Stopped at count =", count - 1, "with total =", total, "\n")
```

## apply Family

```{r apply-family}
# sapply -- simplify result to vector
sapply(1:5, function(x) x^2 + 1)

# lapply -- always returns a list
lapply(1:3, function(x) rep(x, x))

# vapply -- like sapply but with expected output type
vapply(c("hello", "world"), nchar, integer(1))
```

```{r apply-on-matrix}
mat <- matrix(1:12, nrow = 3)
apply(mat, 1, sum)   # row sums
apply(mat, 2, mean)  # column means
```

---

# Writing Functions

```{r simple-function}
greet <- function(name, greeting = "Hello") {
  paste(greeting, name, "- welcome to the workshop!")
}

greet("Amina")
greet("Kofi", greeting = "Bonjour")
```

```{r descriptive-stats-function}
describe <- function(x) {
  list(
    n = length(x),
    mean = mean(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    missing = sum(is.na(x))
  )
}

set.seed(42)
sample_data <- c(rnorm(20, mean = 50, sd = 10), NA, NA)
describe(sample_data)
```

```{r function-returning-df}
generate_data <- function(n = 100, seed = 123) {
  set.seed(seed)
  tibble(
    id = 1:n,
    x = rnorm(n),
    y = 2 * x + rnorm(n, sd = 0.5),
    group = sample(LETTERS[1:3], n, replace = TRUE)
  )
}

sim <- generate_data(10)
sim
```

---

# Pipes

Pipes allow you to chain operations in a readable left-to-right flow.

## Base R Pipe (`|>`)

Available in R >= 4.1.

```{r base-pipe}
# Without pipe
round(mean(c(1.5, 2.7, 3.3, 4.1)), 1)

# With base pipe
c(1.5, 2.7, 3.3, 4.1) |> mean() |> round(1)
```

## magrittr Pipe (`%>%`)

```{r magrittr-pipe}
library(magrittr)

# Equivalent using %>%
c(1.5, 2.7, 3.3, 4.1) %>% mean() %>% round(1)

# Useful with placeholder
c(10, 20, NA, 40) %>% mean(na.rm = TRUE) %>% paste("The mean is", .)
```

```{r pipe-example}
# Practical example: filter and summarise in one pipeline
mtcars |>
  subset(cyl == 6) |>
  (\(d) data.frame(
    n = nrow(d),
    avg_mpg = mean(d$mpg),
    avg_hp = mean(d$hp)
  ))()
```

---

# Lab: Exploring the iris Dataset

The `iris` dataset is built into R and contains measurements of 150 flowers
from three species of iris.

## Exercise 1: Basic Exploration

```{r lab-explore}
# Load the data
data(iris)

# Dimensions
dim(iris)

# Structure
str(iris)

# First and last rows
head(iris)
tail(iris)

# Summary statistics
summary(iris)
```

## Exercise 2: Subsetting

```{r lab-subset}
# Only setosa
setosa <- iris[iris$Species == "setosa", ]
nrow(setosa)

# Flowers with Sepal.Length > 7
large_sepal <- iris[iris$Sepal.Length > 7, ]
large_sepal

# Select only petal columns
petal_data <- iris[, c("Petal.Length", "Petal.Width", "Species")]
head(petal_data)
```

## Exercise 3: Summary Statistics by Species

```{r lab-tapply}
# Mean sepal length by species
tapply(iris$Sepal.Length, iris$Species, mean)

# Standard deviation of petal width by species
tapply(iris$Petal.Width, iris$Species, sd)
```

```{r lab-aggregate}
# aggregate for multiple summaries
aggregate(. ~ Species, data = iris, FUN = mean)
```

## Exercise 4: Custom Function on iris

```{r lab-custom}
species_summary <- function(data, variable) {
  tapply(data[[variable]], data$Species, function(x) {
    c(mean = mean(x), sd = sd(x), min = min(x), max = max(x))
  })
}

species_summary(iris, "Sepal.Length")
species_summary(iris, "Petal.Width")
```

## Exercise 5: Quick Visualisation (Base R)

```{r lab-plot, fig.width=8, fig.height=6}
par(mfrow = c(2, 2))

# Histogram
hist(iris$Sepal.Length, col = "steelblue", main = "Sepal Length Distribution",
     xlab = "Sepal Length (cm)", breaks = 15)

# Boxplot by species
boxplot(Petal.Length ~ Species, data = iris, col = c("coral", "steelblue", "gold"),
        main = "Petal Length by Species", ylab = "Petal Length (cm)")

# Scatter plot
plot(iris$Petal.Length, iris$Petal.Width,
     col = as.numeric(iris$Species) + 1, pch = 19,
     xlab = "Petal Length", ylab = "Petal Width",
     main = "Petal Dimensions")
legend("topleft", levels(iris$Species), col = 2:4, pch = 19, cex = 0.8)

# Pairs plot (subset)
pairs(iris[, 1:4], col = as.numeric(iris$Species) + 1, pch = 19,
      main = "Iris Pairs Plot")

par(mfrow = c(1, 1))
```

---

# Summary

Today we covered:

1. **R basics** -- arithmetic, variables, types
2. **Vectors** -- numeric, character, logical, factor; indexing and operations
3. **Data structures** -- matrices, lists, data frames, tibbles
4. **Control flow** -- if/else, for, while, apply family
5. **Functions** -- writing reusable code
6. **Pipes** -- `|>` and `%>%` for readable pipelines
7. **Lab** -- hands-on exploration of the `iris` dataset

Tomorrow we dive into the **tidyverse** for powerful data wrangling.

```{r session-info}
sessionInfo()
```
