---
title: "Day 2: Tidyverse Data Wrangling"
subtitle: "R for Statistical Analysis Workshop"
author: "Dr. Yae Ulrich Gaba"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Welcome to Day 2. Today we learn the **tidyverse** ecosystem for data
wrangling. The tidyverse is a collection of R packages that share a consistent
design philosophy and grammar. We will focus on `dplyr` for data manipulation,
`tidyr` for reshaping, and `readr` for reading data.

```{r load-packages}
library(tidyverse)  # loads dplyr, tidyr, readr, ggplot2, tibble, etc.
```

---

# Reading Data with readr

## read_csv

`read_csv()` from the readr package is faster and more consistent than base
`read.csv()`. It produces tibbles and does not convert strings to factors.

```{r read-csv-demo}
# Writing a temporary CSV to demonstrate read_csv
temp_file <- tempfile(fileext = ".csv")
writeLines(
  c("name,age,country,score",
    "Amina,24,Nigeria,88",
    "Kofi,27,Ghana,76",
    "Fatou,22,Senegal,92",
    "Emeka,30,Nigeria,85",
    "Ngozi,25,Nigeria,90"),
  temp_file
)

students <- read_csv(temp_file)
students
```

## Creating Sample Data Inline

For workshops and reproducible examples, it is often convenient to create data
directly in R.

```{r inline-data}
# Using tibble()
health_workers <- tibble(
  country = c("Nigeria", "Ghana", "Kenya", "Ethiopia", "South Africa",
              "Senegal", "Tanzania", "Rwanda", "Uganda", "Benin"),
  doctors_per_10k = c(3.8, 1.4, 2.0, 0.8, 9.1,
                      0.7, 0.4, 1.3, 1.7, 1.6),
  nurses_per_10k  = c(14.8, 9.3, 11.7, 7.0, 51.3,
                      4.2, 4.4, 9.5, 13.1, 6.8),
  region = c("West", "West", "East", "East", "Southern",
             "West", "East", "East", "East", "West")
)
health_workers

# Using tribble() for row-by-row entry
gdp_data <- tribble(
  ~country,        ~year, ~gdp_billion_usd,
  "Nigeria",       2020,  432.3,
  "Nigeria",       2021,  440.8,
  "South Africa",  2020,  301.9,
  "South Africa",  2021,  419.0,
  "Kenya",         2020,  100.7,
  "Kenya",         2021,  110.3
)
gdp_data
```

---

# dplyr: Core Verbs

The six core dplyr verbs handle the vast majority of data manipulation tasks.

## filter() -- Select Rows

```{r filter}
# Countries with more than 2 doctors per 10k
health_workers %>%
  filter(doctors_per_10k > 2)

# West African countries with nurses > 8 per 10k
health_workers %>%
  filter(region == "West", nurses_per_10k > 8)

# Multiple conditions with OR
health_workers %>%
  filter(country == "Kenya" | country == "Rwanda")

# Using %in%
health_workers %>%
  filter(country %in% c("Nigeria", "Ghana", "Benin"))
```

## select() -- Choose Columns

```{r select}
# Select specific columns
health_workers %>%
  select(country, doctors_per_10k)

# Drop a column
health_workers %>%
  select(-region)

# Select helpers
health_workers %>%
  select(country, starts_with("doctors"))

health_workers %>%
  select(country, ends_with("per_10k"))

# Rename while selecting
health_workers %>%
  select(nation = country, docs = doctors_per_10k)
```

## mutate() -- Create / Modify Columns

```{r mutate}
health_workers %>%
  mutate(
    total_per_10k = doctors_per_10k + nurses_per_10k,
    doctor_nurse_ratio = round(doctors_per_10k / nurses_per_10k, 2),
    above_avg_docs = doctors_per_10k > mean(doctors_per_10k)
  )
```

```{r mutate-case-when}
# case_when for multiple conditions
health_workers %>%
  mutate(
    doc_level = case_when(
      doctors_per_10k >= 5  ~ "High",
      doctors_per_10k >= 1.5 ~ "Medium",
      TRUE                   ~ "Low"
    )
  )
```

## arrange() -- Sort Rows

```{r arrange}
# Ascending
health_workers %>%
  arrange(doctors_per_10k)

# Descending
health_workers %>%
  arrange(desc(nurses_per_10k))

# Multiple sort keys
health_workers %>%
  arrange(region, desc(doctors_per_10k))
```

## summarise() -- Aggregate

```{r summarise}
health_workers %>%
  summarise(
    n_countries = n(),
    avg_docs = mean(doctors_per_10k),
    sd_docs = sd(doctors_per_10k),
    median_nurses = median(nurses_per_10k),
    min_docs = min(doctors_per_10k),
    max_docs = max(doctors_per_10k)
  )
```

## group_by() + summarise()

```{r group-by}
health_workers %>%
  group_by(region) %>%
  summarise(
    n = n(),
    avg_docs = mean(doctors_per_10k),
    avg_nurses = mean(nurses_per_10k),
    .groups = "drop"
  )
```

```{r group-mutate}
# group_by + mutate for within-group calculations
health_workers %>%
  group_by(region) %>%
  mutate(
    region_avg_docs = mean(doctors_per_10k),
    diff_from_avg = doctors_per_10k - region_avg_docs
  ) %>%
  ungroup()
```

## Chaining Multiple Verbs

```{r chaining}
health_workers %>%
  filter(region %in% c("West", "East")) %>%
  mutate(total_hw = doctors_per_10k + nurses_per_10k) %>%
  group_by(region) %>%
  summarise(
    n = n(),
    avg_total = mean(total_hw),
    best_country = country[which.max(total_hw)],
    .groups = "drop"
  ) %>%
  arrange(desc(avg_total))
```

## Other Useful dplyr Functions

```{r other-dplyr}
# count
health_workers %>% count(region, sort = TRUE)

# slice
health_workers %>% slice_max(doctors_per_10k, n = 3)
health_workers %>% slice_min(nurses_per_10k, n = 3)

# distinct
health_workers %>% distinct(region)

# pull -- extract a column as a vector
health_workers %>% pull(country)

# rename
health_workers %>% rename(nation = country)

# relocate
health_workers %>% relocate(region, .before = country)

# across -- apply function to multiple columns
health_workers %>%
  summarise(across(where(is.numeric), list(mean = mean, sd = sd)))
```

---

# tidyr: Reshaping Data

## Sample Wide Data

```{r wide-data}
# Disease cases by year -- wide format
disease_wide <- tibble(
  country = c("Nigeria", "Ghana", "Kenya", "Ethiopia", "Senegal"),
  cases_2019 = c(12000, 5400, 7800, 9200, 3100),
  cases_2020 = c(15000, 6100, 8900, 10500, 3800),
  cases_2021 = c(13500, 5800, 7200, 11000, 3500),
  cases_2022 = c(11800, 5000, 6800, 10200, 3000)
)
disease_wide
```

## pivot_longer() -- Wide to Long

```{r pivot-longer}
disease_long <- disease_wide %>%
  pivot_longer(
    cols = starts_with("cases_"),
    names_to = "year",
    names_prefix = "cases_",
    values_to = "cases"
  ) %>%
  mutate(year = as.integer(year))

disease_long
```

## pivot_wider() -- Long to Wide

```{r pivot-wider}
# Convert back to wide
disease_long %>%
  pivot_wider(
    names_from = year,
    values_from = cases,
    names_prefix = "cases_"
  )
```

```{r pivot-wider-example}
# Another example: metrics in long format
metrics_long <- tribble(
  ~country,     ~metric,         ~value,
  "Nigeria",    "life_exp",       54.7,
  "Nigeria",    "infant_mort",    72.0,
  "Nigeria",    "literacy",       62.0,
  "Ghana",      "life_exp",       64.1,
  "Ghana",      "infant_mort",    35.0,
  "Ghana",      "literacy",       79.0,
  "Kenya",      "life_exp",       66.7,
  "Kenya",      "infant_mort",    31.0,
  "Kenya",      "literacy",       81.5
)

metrics_long %>%
  pivot_wider(names_from = metric, values_from = value)
```

## separate() and unite()

```{r separate}
# Separate a combined column
records <- tibble(
  id = 1:5,
  date_place = c("2022-03-15_Lagos", "2022-04-20_Accra",
                  "2022-05-10_Nairobi", "2022-06-01_Addis",
                  "2022-07-22_Dakar")
)

records %>%
  separate(date_place, into = c("date", "city"), sep = "_") %>%
  mutate(date = as.Date(date))
```

```{r unite}
# Unite columns
demographics <- tibble(
  first = c("Amina", "Kofi", "Fatou"),
  last = c("Bello", "Mensah", "Diallo"),
  age = c(24, 27, 22)
)

demographics %>%
  unite("full_name", first, last, sep = " ")
```

---

# Joining Data

## Creating Two Tables

```{r join-data}
# Country information
country_info <- tibble(
  country = c("Nigeria", "Ghana", "Kenya", "Ethiopia",
              "South Africa", "Senegal", "Tanzania"),
  population_million = c(211, 31, 54, 115, 60, 17, 62),
  continent_region = c("West Africa", "West Africa", "East Africa",
                       "East Africa", "Southern Africa", "West Africa",
                       "East Africa")
)

# Health spending
health_spending <- tibble(
  country = c("Nigeria", "Ghana", "Kenya", "Ethiopia",
              "Senegal", "Rwanda", "Benin"),
  health_exp_pct_gdp = c(3.9, 3.4, 4.6, 3.5, 3.9, 6.6, 2.6),
  year = rep(2021, 7)
)
```

## left_join()

Keep all rows from the left table; match from the right where possible.

```{r left-join}
country_info %>%
  left_join(health_spending, by = "country")
```

## inner_join()

Keep only rows present in both tables.

```{r inner-join}
country_info %>%
  inner_join(health_spending, by = "country")
```

## Other Joins

```{r other-joins}
# right_join -- keep all rows from right table
country_info %>%
  right_join(health_spending, by = "country")

# full_join -- keep all rows from both
country_info %>%
  full_join(health_spending, by = "country")

# anti_join -- rows in left that have NO match in right
country_info %>%
  anti_join(health_spending, by = "country")

# semi_join -- rows in left that HAVE a match in right (no columns added)
country_info %>%
  semi_join(health_spending, by = "country")
```

---

# Lab: WHO-Style African Health Data

In this lab we create, clean, and analyse an inline dataset that mimics
WHO health indicators for African countries.

## Step 1: Create the Raw Data

```{r lab-raw-data}
# Simulated WHO-style data with some messiness
raw_health <- tibble(
  country = c("Nigeria", "Nigeria", "Nigeria",
              "Ghana", "Ghana", "Ghana",
              "Kenya", "Kenya", "Kenya",
              "Ethiopia", "Ethiopia", "Ethiopia",
              "South Africa", "South Africa", "South Africa",
              "Senegal", "Senegal", "Senegal",
              "Tanzania", "Tanzania", "Tanzania",
              "Rwanda", "Rwanda", "Rwanda"),
  year = rep(c(2018, 2019, 2020), 8),
  life_expectancy = c(54.3, 54.7, 53.9,
                      63.8, 64.1, 63.5,
                      66.3, 66.7, 65.9,
                      65.5, 66.2, 65.0,
                      64.1, 64.7, 63.5,
                      67.5, 67.9, 67.0,
                      65.0, 65.5, 64.8,
                      68.7, 69.0, 68.3),
  infant_mortality_per_1k = c(74.2, 72.0, 75.5,
                               36.2, 35.0, 37.1,
                               32.5, 31.0, 33.8,
                               38.5, 37.0, 40.0,
                               28.3, 27.0, 30.5,
                               33.0, 32.0, 34.5,
                               37.5, 36.0, 39.0,
                               27.8, 26.5, 29.0),
  malaria_cases_thousands = c(58400, 60800, 61700,
                               5200, 5400, 5700,
                               3400, 3600, 3900,
                               2600, 2800, 3000,
                               NA, NA, NA,
                               4800, 5000, 5300,
                               7100, 7400, 7800,
                               1800, 1900, 2100),
  hiv_prevalence_pct = c(1.3, 1.3, 1.3,
                          1.7, 1.7, 1.7,
                          4.5, 4.5, 4.4,
                          0.9, 0.9, 0.9,
                          19.1, 19.0, 18.8,
                          0.4, 0.4, 0.4,
                          4.7, 4.7, 4.6,
                          2.5, 2.5, 2.4),
  region = rep(c("West", "West", "East", "East",
                 "Southern", "West", "East", "East"), each = 3)
)

raw_health
```

## Step 2: Inspect and Clean

```{r lab-inspect}
# Check for missing values
raw_health %>%
  summarise(across(everything(), ~ sum(is.na(.))))

# South Africa has no malaria data -- expected
```

```{r lab-clean}
# Clean: remove NAs for malaria analysis, add a computed column
health_clean <- raw_health %>%
  mutate(
    year = as.integer(year),
    country = factor(country),
    # Health index: higher life expectancy + lower infant mortality = better
    health_index = (life_expectancy / 100) - (infant_mortality_per_1k / 200)
  )

health_clean
```

## Step 3: Summary by Region and Year

```{r lab-regional}
regional_summary <- health_clean %>%
  group_by(region, year) %>%
  summarise(
    n_countries = n_distinct(country),
    avg_life_exp = round(mean(life_expectancy), 1),
    avg_infant_mort = round(mean(infant_mortality_per_1k), 1),
    total_malaria_k = sum(malaria_cases_thousands, na.rm = TRUE),
    .groups = "drop"
  )

regional_summary
```

## Step 4: Identify Trends

```{r lab-trends}
# Year-over-year change in life expectancy
trends <- health_clean %>%
  arrange(country, year) %>%
  group_by(country) %>%
  mutate(
    le_change = life_expectancy - lag(life_expectancy),
    im_change = infant_mortality_per_1k - lag(infant_mortality_per_1k)
  ) %>%
  ungroup() %>%
  filter(!is.na(le_change))

# Which countries improved in 2019 vs 2018?
trends %>%
  filter(year == 2019, le_change > 0) %>%
  select(country, year, life_expectancy, le_change) %>%
  arrange(desc(le_change))
```

## Step 5: Reshape for Reporting

```{r lab-reshape}
# Create a wide table of life expectancy by year
le_wide <- health_clean %>%
  select(country, year, life_expectancy) %>%
  pivot_wider(
    names_from = year,
    values_from = life_expectancy,
    names_prefix = "le_"
  ) %>%
  mutate(change_2018_2020 = le_2020 - le_2018)

le_wide
```

## Step 6: Join with External Data

```{r lab-join}
# GDP per capita (simulated)
gdp_pc <- tibble(
  country = c("Nigeria", "Ghana", "Kenya", "Ethiopia",
              "South Africa", "Senegal", "Tanzania", "Rwanda"),
  gdp_per_capita = c(2097, 2328, 1838, 936, 5091, 1488, 1076, 822)
)

# Join and correlate
combined <- health_clean %>%
  filter(year == 2020) %>%
  inner_join(gdp_pc, by = "country")

combined %>%
  select(country, life_expectancy, infant_mortality_per_1k, gdp_per_capita) %>%
  arrange(desc(gdp_per_capita))

# Quick correlation
cor(combined$gdp_per_capita, combined$life_expectancy)
cor(combined$gdp_per_capita, combined$infant_mortality_per_1k)
```

---

# Summary

Today we covered:

1. **readr** -- `read_csv()` for fast, consistent data import
2. **dplyr verbs** -- `filter`, `select`, `mutate`, `arrange`, `summarise`, `group_by`
3. **tidyr** -- `pivot_longer`, `pivot_wider`, `separate`, `unite`
4. **Joins** -- `left_join`, `inner_join`, `right_join`, `full_join`, `anti_join`
5. **Lab** -- end-to-end wrangling of WHO-style health data for African countries

Tomorrow we move to **ggplot2** for beautiful, publication-quality data
visualisation.

```{r session-info}
sessionInfo()
```
