---
title: "Day 3: Data Visualisation with ggplot2"
subtitle: "R for Statistical Analysis Workshop"
author: "Dr. Yae Ulrich Gaba"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150
)
```

# Introduction

Today we master **ggplot2**, the most popular visualisation package in R. It
implements the *Grammar of Graphics*: every plot is built from layers that
combine data, aesthetic mappings, geometric objects, and scales.

```{r load-packages}
library(tidyverse)
library(patchwork)   # for combining plots
library(scales)      # for axis formatting
```

## Prepare Data

We will reuse our African health data and also use built-in datasets.

```{r data-prep}
health <- tibble(
  country = rep(c("Nigeria", "Ghana", "Kenya", "Ethiopia",
                  "South Africa", "Senegal", "Tanzania", "Rwanda"), each = 3),
  year = rep(c(2018, 2019, 2020), 8),
  life_expectancy = c(54.3, 54.7, 53.9,  63.8, 64.1, 63.5,
                      66.3, 66.7, 65.9,  65.5, 66.2, 65.0,
                      64.1, 64.7, 63.5,  67.5, 67.9, 67.0,
                      65.0, 65.5, 64.8,  68.7, 69.0, 68.3),
  infant_mortality = c(74.2, 72.0, 75.5,  36.2, 35.0, 37.1,
                       32.5, 31.0, 33.8,  38.5, 37.0, 40.0,
                       28.3, 27.0, 30.5,  33.0, 32.0, 34.5,
                       37.5, 36.0, 39.0,  27.8, 26.5, 29.0),
  region = rep(c("West", "West", "East", "East",
                 "Southern", "West", "East", "East"), each = 3),
  gdp_per_capita = rep(c(2097, 2328, 1838, 936,
                          5091, 1488, 1076, 822), each = 3),
  population_m = rep(c(211, 31, 54, 115, 60, 17, 62, 13), each = 3)
)

health_2020 <- health %>% filter(year == 2020)
```

---

# The Grammar of Graphics

A ggplot is built with these components:

| Component | Function | Purpose |
|-----------|----------|---------|
| Data | `ggplot(data)` | The dataset |
| Aesthetics | `aes(x, y, color, ...)` | Map variables to visual properties |
| Geometries | `geom_*()` | How to draw the data |
| Facets | `facet_wrap()` / `facet_grid()` | Small multiples |
| Scales | `scale_*()` | Control mappings (colours, axes) |
| Theme | `theme_*()` | Non-data visual elements |

```{r grammar-minimal}
# Minimal ggplot
ggplot(data = health_2020, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point()
```

---

# Scatter Plots with geom_point()

```{r scatter-basic}
ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(size = 3) +
  labs(title = "Life Expectancy vs GDP per Capita (2020)",
       x = "GDP per Capita (USD)",
       y = "Life Expectancy (years)")
```

```{r scatter-enhanced}
ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy,
                         color = region, size = population_m)) +
  geom_point(alpha = 0.8) +
  geom_text(aes(label = country), vjust = -1, size = 3, show.legend = FALSE) +
  scale_size_continuous(name = "Population (M)", range = c(3, 12)) +
  scale_color_brewer(palette = "Set2", name = "Region") +
  labs(title = "Life Expectancy vs GDP per Capita",
       subtitle = "African countries, 2020",
       x = "GDP per Capita (USD)",
       y = "Life Expectancy (years)") +
  theme_minimal(base_size = 13)
```

---

# Line Plots with geom_line()

```{r line-basic}
ggplot(health, aes(x = year, y = life_expectancy, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Life Expectancy Trends (2018-2020)",
       x = "Year", y = "Life Expectancy (years)",
       color = "Country") +
  scale_x_continuous(breaks = 2018:2020) +
  theme_minimal(base_size = 13)
```

```{r line-regional}
regional_avg <- health %>%
  group_by(region, year) %>%
  summarise(avg_le = mean(life_expectancy), .groups = "drop")

ggplot(regional_avg, aes(x = year, y = avg_le, color = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2018:2020) +
  scale_color_manual(values = c("East" = "#2C7BB6", "Southern" = "#D7191C",
                                 "West" = "#FDAE61")) +
  labs(title = "Average Life Expectancy by Region",
       x = "Year", y = "Avg Life Expectancy", color = "Region") +
  theme_light(base_size = 13)
```

---

# Bar Charts with geom_bar() and geom_col()

`geom_bar()` counts observations; `geom_col()` uses a pre-computed y value.

```{r bar-col}
ggplot(health_2020, aes(x = reorder(country, life_expectancy),
                         y = life_expectancy, fill = region)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Life Expectancy by Country (2020)",
       x = NULL, y = "Life Expectancy (years)", fill = "Region") +
  theme_minimal(base_size = 13)
```

```{r bar-grouped}
# Grouped bar chart: infant mortality across years
health_subset <- health %>%
  filter(country %in% c("Nigeria", "Ghana", "Kenya", "Rwanda"))

ggplot(health_subset, aes(x = country, y = infant_mortality,
                           fill = factor(year))) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_viridis_d(name = "Year", option = "D", begin = 0.3, end = 0.9) +
  labs(title = "Infant Mortality by Country and Year",
       x = NULL, y = "Infant Mortality (per 1,000)") +
  theme_minimal(base_size = 13)
```

---

# Histograms with geom_histogram()

```{r histogram}
set.seed(42)
exam_scores <- tibble(
  score = c(rnorm(120, mean = 65, sd = 12),
            rnorm(80, mean = 78, sd = 8)),
  group = c(rep("Group A", 120), rep("Group B", 80))
)

ggplot(exam_scores, aes(x = score)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white",
                 alpha = 0.8) +
  labs(title = "Distribution of Exam Scores",
       x = "Score", y = "Count") +
  theme_minimal(base_size = 13)
```

```{r histogram-grouped}
ggplot(exam_scores, aes(x = score, fill = group)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity",
                 color = "white") +
  scale_fill_manual(values = c("Group A" = "#E69F00", "Group B" = "#56B4E9")) +
  labs(title = "Score Distributions by Group",
       x = "Score", y = "Count", fill = "Group") +
  theme_minimal(base_size = 13)
```

---

# Density Plots with geom_density()

```{r density}
ggplot(exam_scores, aes(x = score, fill = group, color = group)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
  scale_color_manual(values = c("#E69F00", "#56B4E9")) +
  labs(title = "Score Density by Group",
       x = "Score", y = "Density") +
  theme_minimal(base_size = 13)
```

---

# Box Plots with geom_boxplot()

```{r boxplot}
ggplot(health, aes(x = region, y = life_expectancy, fill = region)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Life Expectancy Distribution by Region",
       x = "Region", y = "Life Expectancy (years)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r boxplot-year}
ggplot(health, aes(x = factor(year), y = infant_mortality, fill = factor(year))) +
  geom_boxplot(alpha = 0.6) +
  scale_fill_viridis_d(option = "C", begin = 0.3, end = 0.9) +
  labs(title = "Infant Mortality Distribution by Year",
       x = "Year", y = "Infant Mortality (per 1,000)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# Customisation

## Themes

```{r themes, fig.height=8}
base <- ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(size = 3, color = "steelblue") +
  labs(title = "Theme Comparison", x = "GDP/capita", y = "Life Exp.")

p1 <- base + theme_gray() + ggtitle("theme_gray (default)")
p2 <- base + theme_minimal() + ggtitle("theme_minimal")
p3 <- base + theme_classic() + ggtitle("theme_classic")
p4 <- base + theme_dark() + ggtitle("theme_dark")

(p1 | p2) / (p3 | p4)
```

## Colour Scales

```{r colour-scales}
# Continuous colour
ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy,
                         color = infant_mortality)) +
  geom_point(size = 5) +
  scale_color_viridis_c(option = "plasma", name = "Infant\nMortality") +
  labs(title = "Colour by Infant Mortality",
       x = "GDP per Capita", y = "Life Expectancy") +
  theme_minimal(base_size = 13)
```

## Labels and Annotations

```{r annotations}
ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(aes(color = region), size = 4) +
  geom_smooth(method = "lm", se = TRUE, color = "gray40", linetype = "dashed") +
  annotate("text", x = 4000, y = 55, label = "Nigeria stands out\nwith low life expectancy",
           hjust = 0, size = 3.5, color = "gray30", fontface = "italic") +
  annotate("rect", xmin = 4500, xmax = 5500, ymin = 62.5, ymax = 64.5,
           alpha = 0.15, fill = "red") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Life Expectancy vs GDP per Capita",
       subtitle = "With linear trend and annotations",
       x = "GDP per Capita (USD)",
       y = "Life Expectancy (years)",
       color = "Region",
       caption = "Source: Simulated WHO-style data") +
  theme_minimal(base_size = 13)
```

---

# Faceting

## facet_wrap()

```{r facet-wrap}
ggplot(health, aes(x = year, y = life_expectancy)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  facet_wrap(~ country, ncol = 4, scales = "free_y") +
  scale_x_continuous(breaks = 2018:2020) +
  labs(title = "Life Expectancy Trends by Country",
       x = "Year", y = "Life Expectancy") +
  theme_light(base_size = 11) +
  theme(strip.background = element_rect(fill = "steelblue"),
        strip.text = element_text(color = "white", face = "bold"))
```

## facet_grid()

```{r facet-grid, fig.height=7}
# Create a factor for better ordering
health <- health %>%
  mutate(region_f = factor(region, levels = c("West", "East", "Southern")))

ggplot(health, aes(x = year, y = infant_mortality, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_grid(region_f ~ ., scales = "free_y") +
  scale_x_continuous(breaks = 2018:2020) +
  labs(title = "Infant Mortality Trends",
       subtitle = "Faceted by region",
       x = "Year", y = "Infant Mortality (per 1,000)",
       color = "Country") +
  theme_minimal(base_size = 12) +
  theme(strip.text.y = element_text(angle = 0, face = "bold"))
```

---

# Combining Plots with patchwork

The `patchwork` package makes it easy to compose multiple ggplots.

```{r patchwork-demo, fig.width=10, fig.height=8}
p1 <- ggplot(health_2020, aes(x = reorder(country, life_expectancy),
                                y = life_expectancy)) +
  geom_col(fill = "steelblue", width = 0.7) +
  coord_flip() +
  labs(title = "A) Life Expectancy", x = NULL, y = "Years") +
  theme_minimal(base_size = 11)

p2 <- ggplot(health_2020, aes(x = reorder(country, -infant_mortality),
                                y = infant_mortality)) +
  geom_col(fill = "coral", width = 0.7) +
  coord_flip() +
  labs(title = "B) Infant Mortality", x = NULL, y = "Per 1,000 births") +
  theme_minimal(base_size = 11)

p3 <- ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy,
                                color = region)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "gray50",
              linetype = "dashed") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "C) GDP vs Life Expectancy",
       x = "GDP per Capita", y = "Life Exp.") +
  theme_minimal(base_size = 11)

p4 <- ggplot(health_2020, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point(aes(size = population_m), alpha = 0.7, color = "purple") +
  geom_smooth(method = "lm", se = TRUE, color = "gray40") +
  scale_size_continuous(name = "Pop. (M)") +
  labs(title = "D) Infant Mortality vs Life Expectancy",
       x = "Infant Mortality", y = "Life Exp.") +
  theme_minimal(base_size = 11)

# Compose
(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title = "Health Indicators Across African Countries (2020)",
    subtitle = "A multi-panel overview",
    caption = "Source: Simulated WHO-style data",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )
```

---

# Lab: Create 6+ Publication-Quality Figures

## Figure 1: Lollipop Chart

```{r lab-fig1, fig.width=8, fig.height=5}
ggplot(health_2020, aes(x = reorder(country, life_expectancy),
                         y = life_expectancy)) +
  geom_segment(aes(xend = country, y = 50, yend = life_expectancy),
               color = "gray60", linewidth = 0.8) +
  geom_point(aes(color = region), size = 5) +
  coord_flip() +
  scale_color_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(50, 72)) +
  labs(title = "Life Expectancy in African Countries (2020)",
       subtitle = "Lollipop chart ordered by life expectancy",
       x = NULL, y = "Life Expectancy (years)", color = "Region") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

## Figure 2: Heatmap

```{r lab-fig2, fig.width=9, fig.height=5}
health_scaled <- health %>%
  group_by(year) %>%
  mutate(le_zscore = (life_expectancy - mean(life_expectancy)) / sd(life_expectancy)) %>%
  ungroup()

ggplot(health_scaled, aes(x = factor(year), y = reorder(country, le_zscore),
                           fill = le_zscore)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(low = "#d73027", mid = "#ffffbf", high = "#1a9850",
                       midpoint = 0, name = "Z-score") +
  geom_text(aes(label = round(life_expectancy, 1)), size = 3.5) +
  labs(title = "Life Expectancy Heatmap",
       subtitle = "Colour indicates relative standing (z-score) within each year",
       x = "Year", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

## Figure 3: Dumbbell Chart (Change from 2018 to 2020)

```{r lab-fig3, fig.width=8, fig.height=5}
change_data <- health %>%
  filter(year %in% c(2018, 2020)) %>%
  select(country, year, life_expectancy) %>%
  pivot_wider(names_from = year, values_from = life_expectancy,
              names_prefix = "le_")

ggplot(change_data, aes(y = reorder(country, le_2020))) +
  geom_segment(aes(x = le_2018, xend = le_2020, yend = country),
               color = "gray60", linewidth = 1) +
  geom_point(aes(x = le_2018), color = "#E69F00", size = 4) +
  geom_point(aes(x = le_2020), color = "#0072B2", size = 4) +
  annotate("text", x = 56, y = 8.3, label = "2018", color = "#E69F00",
           fontface = "bold", size = 4) +
  annotate("text", x = 57.5, y = 8.3, label = "2020", color = "#0072B2",
           fontface = "bold", size = 4) +
  labs(title = "Change in Life Expectancy (2018 vs 2020)",
       subtitle = "Most countries saw a decline in 2020",
       x = "Life Expectancy (years)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

## Figure 4: Scatter with Marginal Density (using cowplot approach)

```{r lab-fig4, fig.width=8, fig.height=7}
main_plot <- ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy,
                                      color = region)) +
  geom_point(size = 5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(x = "GDP per Capita (USD)", y = "Life Expectancy (years)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

top_density <- ggplot(health_2020, aes(x = gdp_per_capita, fill = region)) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette = "Set1") +
  theme_void() +
  theme(legend.position = "none")

right_density <- ggplot(health_2020, aes(x = life_expectancy, fill = region)) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette = "Set1") +
  coord_flip() +
  theme_void() +
  theme(legend.position = "none")

top_density + plot_spacer() + main_plot + right_density +
  plot_layout(ncol = 2, widths = c(4, 1), heights = c(1, 4)) +
  plot_annotation(title = "GDP vs Life Expectancy with Marginal Densities")
```

## Figure 5: Stacked Area Chart

```{r lab-fig5, fig.width=9, fig.height=5}
malaria_data <- tibble(
  country = rep(c("Nigeria", "Ghana", "Kenya", "Ethiopia",
                  "Senegal", "Tanzania"), each = 5),
  year = rep(2016:2020, 6),
  cases_k = c(55000, 56500, 58400, 60800, 61700,
              4800, 5000, 5200, 5400, 5700,
              3000, 3200, 3400, 3600, 3900,
              2200, 2400, 2600, 2800, 3000,
              4200, 4500, 4800, 5000, 5300,
              6500, 6800, 7100, 7400, 7800)
)

ggplot(malaria_data %>% filter(country != "Nigeria"),
       aes(x = year, y = cases_k / 1000, fill = country)) +
  geom_area(alpha = 0.8, color = "white", linewidth = 0.3) +
  scale_fill_viridis_d(option = "turbo", begin = 0.1, end = 0.9) +
  scale_x_continuous(breaks = 2016:2020) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Malaria Cases in Selected African Countries",
       subtitle = "Nigeria excluded due to scale (>55M cases)",
       x = "Year", y = "Cases (thousands)", fill = "Country") +
  theme_minimal(base_size = 13)
```

## Figure 6: Diverging Bar Chart

```{r lab-fig6, fig.width=8, fig.height=5}
avg_le <- mean(health_2020$life_expectancy)

diverging <- health_2020 %>%
  mutate(
    diff = life_expectancy - avg_le,
    above = diff > 0
  )

ggplot(diverging, aes(x = reorder(country, diff), y = diff, fill = above)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2C7BB6", "FALSE" = "#D7191C")) +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  labs(title = "Deviation from Average Life Expectancy (2020)",
       subtitle = paste0("Average = ", round(avg_le, 1), " years"),
       x = NULL, y = "Difference from average (years)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

## Figure 7 (Bonus): Ridgeline-Style Density with facet

```{r lab-fig7, fig.width=8, fig.height=6}
set.seed(123)
sim_scores <- tibble(
  country = rep(c("Nigeria", "Ghana", "Kenya", "Rwanda",
                  "South Africa", "Senegal"), each = 200),
  score = c(rnorm(200, 52, 15), rnorm(200, 60, 12), rnorm(200, 65, 10),
            rnorm(200, 70, 8), rnorm(200, 63, 14), rnorm(200, 58, 11))
)

ggplot(sim_scores, aes(x = score, fill = country, color = country)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ country, ncol = 2, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  scale_color_viridis_d(option = "D") +
  labs(title = "Simulated Exam Score Distributions by Country",
       x = "Score", y = "Density") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))
```

---

# Summary

Today we covered:

1. **Grammar of Graphics** -- data, aesthetics, geometries, scales, themes
2. **Core geoms** -- `geom_point`, `geom_line`, `geom_bar`/`geom_col`, `geom_histogram`, `geom_boxplot`, `geom_density`
3. **Customisation** -- themes, colour scales, labels, annotations
4. **Faceting** -- `facet_wrap` and `facet_grid`
5. **patchwork** -- composing multiple plots
6. **Lab** -- 7 publication-quality figures (lollipop, heatmap, dumbbell, scatter with marginals, stacked area, diverging bar, faceted density)

Tomorrow we bring it all together with **statistical modelling** and
**reproducible reporting** in R Markdown.

```{r session-info}
sessionInfo()
```
