---
title: "Day 4: Statistical Modelling & Reproducible Reporting"
subtitle: "R for Statistical Analysis Workshop"
author: "Dr. Yae Ulrich Gaba"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150
)
```

# Introduction

On this final day we cover **statistical modelling** in R -- from descriptive
statistics through linear and logistic regression -- and learn how to turn
analyses into **reproducible reports** using R Markdown features.

```{r load-packages}
library(tidyverse)
library(broom)       # tidy model output
library(knitr)       # kable tables
library(patchwork)
```

## Prepare Data

```{r data}
set.seed(42)

# African health dataset
health <- tibble(
  country = rep(c("Nigeria", "Ghana", "Kenya", "Ethiopia",
                  "South Africa", "Senegal", "Tanzania", "Rwanda",
                  "Uganda", "Benin"), each = 3),
  year = rep(2018:2020, 10),
  life_expectancy = c(54.3, 54.7, 53.9,  63.8, 64.1, 63.5,
                      66.3, 66.7, 65.9,  65.5, 66.2, 65.0,
                      64.1, 64.7, 63.5,  67.5, 67.9, 67.0,
                      65.0, 65.5, 64.8,  68.7, 69.0, 68.3,
                      63.0, 63.5, 62.8,  61.5, 61.8, 61.0),
  infant_mortality = c(74.2, 72.0, 75.5,  36.2, 35.0, 37.1,
                       32.5, 31.0, 33.8,  38.5, 37.0, 40.0,
                       28.3, 27.0, 30.5,  33.0, 32.0, 34.5,
                       37.5, 36.0, 39.0,  27.8, 26.5, 29.0,
                       41.0, 39.5, 43.0,  55.0, 53.5, 57.0),
  gdp_per_capita = rep(c(2097, 2328, 1838, 936, 5091,
                          1488, 1076, 822, 817, 1291), each = 3),
  health_exp_pct = rep(c(3.9, 3.4, 4.6, 3.5, 8.3,
                          3.9, 3.6, 6.6, 3.8, 2.6), each = 3),
  region = rep(c("West", "West", "East", "East", "Southern",
                 "West", "East", "East", "East", "West"), each = 3)
)

health_2020 <- health %>% filter(year == 2020)
```

---

# Descriptive Statistics

## Numerical Summaries

```{r descriptive}
# Full summary
summary(health_2020[, c("life_expectancy", "infant_mortality",
                         "gdp_per_capita", "health_exp_pct")])
```

```{r descriptive-custom}
# Custom summary table
desc_table <- health_2020 %>%
  summarise(
    across(
      c(life_expectancy, infant_mortality, gdp_per_capita),
      list(
        mean = ~ mean(.x),
        sd = ~ sd(.x),
        median = ~ median(.x),
        min = ~ min(.x),
        max = ~ max(.x)
      ),
      .names = "{.col}__{.fn}"
    )
  ) %>%
  pivot_longer(everything(),
               names_to = c("variable", "statistic"),
               names_sep = "__",
               values_to = "value") %>%
  pivot_wider(names_from = statistic, values_from = value)

kable(desc_table, digits = 2, caption = "Descriptive Statistics (2020)")
```

## Group Summaries

```{r group-desc}
health_2020 %>%
  group_by(region) %>%
  summarise(
    n = n(),
    mean_le = mean(life_expectancy),
    sd_le = sd(life_expectancy),
    mean_im = mean(infant_mortality),
    .groups = "drop"
  ) %>%
  kable(digits = 2, caption = "Summary by Region")
```

---

# Hypothesis Tests

## Two-Sample t-test

We test whether life expectancy differs between West and East African countries.

```{r t-test}
west <- health_2020 %>% filter(region == "West") %>% pull(life_expectancy)
east <- health_2020 %>% filter(region == "East") %>% pull(life_expectancy)

t_result <- t.test(west, east, alternative = "two.sided")
t_result
```

```{r t-test-tidy}
# Tidy output with broom
tidy(t_result) %>%
  kable(digits = 3, caption = "Two-sample t-test: West vs East Life Expectancy")
```

**Interpretation:** The p-value is `r round(t_result$p.value, 4)`. At the 0.05
significance level, we
`r ifelse(t_result$p.value < 0.05, "reject", "fail to reject")` the null
hypothesis of equal means.

## Paired t-test

Compare 2018 vs 2020 life expectancy within countries.

```{r paired-t}
le_2018 <- health %>% filter(year == 2018) %>% pull(life_expectancy)
le_2020 <- health %>% filter(year == 2020) %>% pull(life_expectancy)

paired_result <- t.test(le_2018, le_2020, paired = TRUE)
tidy(paired_result) %>%
  kable(digits = 3, caption = "Paired t-test: 2018 vs 2020 Life Expectancy")
```

## Chi-Squared Test

Test whether there is an association between region and a binary health
indicator.

```{r chi-sq}
health_2020 <- health_2020 %>%
  mutate(high_le = ifelse(life_expectancy > median(life_expectancy),
                           "Above Median", "Below Median"))

# Contingency table
ct <- table(health_2020$region, health_2020$high_le)
ct

chi_result <- chisq.test(ct, simulate.p.value = TRUE, B = 5000)
chi_result

tidy(chi_result) %>%
  kable(digits = 3, caption = "Chi-squared Test: Region vs Life Expectancy Category")
```

---

# Simple Linear Regression

We model life expectancy as a function of GDP per capita.

```{r slr}
model1 <- lm(life_expectancy ~ gdp_per_capita, data = health_2020)
summary(model1)
```

```{r slr-tidy}
# Tidy coefficients
tidy(model1, conf.int = TRUE) %>%
  kable(digits = 4, caption = "Simple Linear Regression Coefficients")

# Model-level summary
glance(model1) %>%
  kable(digits = 3, caption = "Model Fit Statistics")
```

## Visualise the Fit

```{r slr-plot}
ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(aes(color = region), size = 4) +
  geom_smooth(method = "lm", se = TRUE, color = "gray30") +
  geom_text(aes(label = country), vjust = -1, size = 3) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Life Expectancy ~ GDP per Capita",
       subtitle = paste0("R-squared = ", round(summary(model1)$r.squared, 3)),
       x = "GDP per Capita (USD)", y = "Life Expectancy (years)") +
  theme_minimal(base_size = 13)
```

## Residual Diagnostics

```{r residual-diag, fig.height=8}
par(mfrow = c(2, 2))
plot(model1)
par(mfrow = c(1, 1))
```

```{r residual-tidy}
# Augmented data with residuals and fitted values
augmented1 <- augment(model1)
head(augmented1) %>% kable(digits = 3)
```

```{r residual-ggplot}
p_resid <- ggplot(augmented1, aes(x = .fitted, y = .resid)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, color = "steelblue") +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

p_qq <- ggplot(augmented1, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot") +
  theme_minimal()

p_resid | p_qq
```

---

# Multiple Regression

Now we add infant mortality and health expenditure as predictors.

```{r multiple-reg}
model2 <- lm(life_expectancy ~ gdp_per_capita + infant_mortality + health_exp_pct,
             data = health_2020)
summary(model2)
```

```{r multiple-reg-tidy}
tidy(model2, conf.int = TRUE) %>%
  kable(digits = 4, caption = "Multiple Regression Coefficients")

glance(model2) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC) %>%
  kable(digits = 3, caption = "Multiple Regression Fit")
```

## Comparing Models

```{r compare-models}
# Compare R-squared
tibble(
  model = c("Simple (GDP only)", "Multiple (GDP + IM + Health Exp)"),
  r_squared = c(summary(model1)$r.squared, summary(model2)$r.squared),
  adj_r_squared = c(summary(model1)$adj.r.squared, summary(model2)$adj.r.squared),
  AIC = c(AIC(model1), AIC(model2)),
  BIC = c(BIC(model1), BIC(model2))
) %>%
  kable(digits = 3, caption = "Model Comparison")

# F-test for nested models
anova(model1, model2)
```

---

# ANOVA

We use one-way ANOVA to test whether life expectancy differs across regions.

```{r anova}
# Using the full dataset for more observations
aov_model <- aov(life_expectancy ~ region, data = health)
summary(aov_model)
```

```{r anova-tidy}
tidy(aov_model) %>%
  kable(digits = 3, caption = "One-way ANOVA: Life Expectancy by Region")
```

```{r tukey}
# Post-hoc pairwise comparisons
tukey <- TukeyHSD(aov_model)
tukey

tidy(tukey) %>%
  kable(digits = 3, caption = "Tukey HSD Pairwise Comparisons")
```

```{r anova-plot}
ggplot(health, aes(x = region, y = life_expectancy, fill = region)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "ANOVA: Life Expectancy by Region",
       subtitle = paste0("F = ", round(summary(aov_model)[[1]][["F value"]][1], 2),
                         ", p = ", format.pval(summary(aov_model)[[1]][["Pr(>F)"]][1],
                                               digits = 3)),
       x = "Region", y = "Life Expectancy (years)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# Logistic Regression

We create a binary outcome: whether a country achieves "good" life expectancy
(above 65 years).

```{r logistic-data}
health_logit <- health %>%
  mutate(good_le = as.integer(life_expectancy > 65))

table(health_logit$good_le)
```

```{r logistic-model}
logit_model <- glm(good_le ~ gdp_per_capita + infant_mortality + health_exp_pct,
                   data = health_logit, family = binomial(link = "logit"))
summary(logit_model)
```

```{r logistic-tidy}
tidy(logit_model, conf.int = TRUE, exponentiate = TRUE) %>%
  kable(digits = 3, caption = "Logistic Regression: Odds Ratios")
```

```{r logistic-interpret}
# Model fit
glance(logit_model) %>%
  kable(digits = 3, caption = "Logistic Regression Fit Statistics")
```

```{r logistic-predict}
# Predicted probabilities
health_logit <- health_logit %>%
  mutate(pred_prob = predict(logit_model, type = "response"),
         pred_class = ifelse(pred_prob > 0.5, 1, 0))

# Confusion matrix
table(Predicted = health_logit$pred_class, Actual = health_logit$good_le) %>%
  kable(caption = "Confusion Matrix")

# Accuracy
accuracy <- mean(health_logit$pred_class == health_logit$good_le)
cat("Accuracy:", round(accuracy * 100, 1), "%\n")
```

```{r logistic-plot}
ggplot(health_logit, aes(x = infant_mortality, y = pred_prob,
                          color = factor(good_le))) +
  geom_point(size = 3, alpha = 0.7) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("0" = "#D7191C", "1" = "#1A9850"),
                     labels = c("LE <= 65", "LE > 65"),
                     name = "Actual") +
  labs(title = "Predicted Probability of Good Life Expectancy",
       x = "Infant Mortality (per 1,000)", y = "Predicted Probability") +
  theme_minimal(base_size = 13)
```

---

# Tidy Modelling with broom

The `broom` package provides three key functions for working with model objects.

```{r broom-overview}
# tidy() -- coefficient-level summary
tidy(model2) %>% kable(digits = 4)

# glance() -- model-level summary
glance(model2) %>% kable(digits = 3)

# augment() -- observation-level data (fitted, residuals, etc.)
augment(model2) %>%
  head(5) %>%
  kable(digits = 3)
```

```{r broom-multiple-models}
# Fit models for each region and tidy the output
health %>%
  group_by(region) %>%
  group_modify(~ {
    mod <- lm(life_expectancy ~ infant_mortality, data = .x)
    tidy(mod, conf.int = TRUE)
  }) %>%
  filter(term != "(Intercept)") %>%
  kable(digits = 3, caption = "Infant Mortality Effect on Life Expectancy by Region")
```

---

# R Markdown Features

This entire document is itself an R Markdown report. Here we highlight key
features.

## Code Chunk Options

Code chunks accept options that control their behaviour:

- `echo = FALSE` -- hide code, show output
- `eval = FALSE` -- show code, do not run
- `include = FALSE` -- run code but show neither code nor output
- `fig.width`, `fig.height`, `fig.cap` -- figure sizing and captions
- `cache = TRUE` -- cache results for faster re-knitting

## Inline Code

You can embed R results directly in text. For example:

The dataset has `r nrow(health)` observations across
`r n_distinct(health$country)` countries. The mean life expectancy in 2020 was
`r round(mean(health_2020$life_expectancy), 1)` years.

This is written as:

````
The dataset has `r knitr::inline_expr("nrow(health)")` observations across
`r knitr::inline_expr("n_distinct(health$country)")` countries.
````

## kable Tables

The `knitr::kable()` function creates clean tables for HTML and PDF output.

```{r kable-demo}
health_2020 %>%
  select(country, region, life_expectancy, infant_mortality, gdp_per_capita) %>%
  arrange(desc(life_expectancy)) %>%
  kable(digits = 1,
        col.names = c("Country", "Region", "Life Exp.", "Infant Mort.",
                       "GDP/capita"),
        caption = "Health Indicators by Country (2020)",
        format.args = list(big.mark = ","))
```

## Cross-Referencing Values

When writing a report, refer to computed values rather than hard-coding numbers:

```{r compute-for-report}
model_r2 <- round(summary(model2)$r.squared * 100, 1)
best_country <- health_2020 %>%
  slice_max(life_expectancy) %>%
  pull(country)
worst_country <- health_2020 %>%
  slice_min(life_expectancy) %>%
  pull(country)
```

Our multiple regression model explains `r model_r2`% of the variance in life
expectancy. `r best_country` had the highest life expectancy, while
`r worst_country` had the lowest.

---

# Lab: Complete Reproducible Report

Below is a self-contained mini-report that demonstrates the full workflow:
data, analysis, visualisation, and narrative.

## Research Question

**Does health expenditure as a percentage of GDP predict life expectancy in
African countries, after controlling for GDP per capita?**

## Data Overview

```{r lab-overview}
health_2020 %>%
  select(country, region, life_expectancy, gdp_per_capita, health_exp_pct) %>%
  kable(digits = 1, caption = "Study Data (2020)")
```

We have data on `r nrow(health_2020)` African countries. The mean life
expectancy is `r round(mean(health_2020$life_expectancy), 1)` years (SD =
`r round(sd(health_2020$life_expectancy), 1)`).

## Exploratory Analysis

```{r lab-explore, fig.height=4}
p1 <- ggplot(health_2020, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(aes(color = region), size = 4) +
  geom_smooth(method = "lm", se = TRUE, color = "gray40") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "GDP vs Life Expectancy", x = "GDP/capita", y = "Life Exp.") +
  theme_minimal()

p2 <- ggplot(health_2020, aes(x = health_exp_pct, y = life_expectancy)) +
  geom_point(aes(color = region), size = 4) +
  geom_smooth(method = "lm", se = TRUE, color = "gray40") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Health Exp. vs Life Expectancy", x = "Health Exp. (% GDP)",
       y = "Life Exp.") +
  theme_minimal()

p1 + p2 + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## Correlation Matrix

```{r lab-corr}
health_2020 %>%
  select(life_expectancy, gdp_per_capita, health_exp_pct, infant_mortality) %>%
  cor() %>%
  round(3) %>%
  kable(caption = "Correlation Matrix")
```

## Statistical Model

```{r lab-model}
report_model <- lm(life_expectancy ~ gdp_per_capita + health_exp_pct,
                   data = health_2020)

tidy(report_model, conf.int = TRUE) %>%
  kable(digits = 4, caption = "Regression Results")

glance(report_model) %>%
  select(r.squared, adj.r.squared, sigma, p.value, AIC) %>%
  kable(digits = 3, caption = "Model Fit")
```

```{r lab-diagnostics, fig.height=4}
aug <- augment(report_model)

p_resid <- ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted", x = "Fitted", y = "Residuals") +
  theme_minimal()

p_qq <- ggplot(aug, aes(sample = .resid)) +
  stat_qq() + stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot") +
  theme_minimal()

p_resid | p_qq
```

## Conclusions

```{r lab-conclusions}
coefs <- tidy(report_model)
gdp_coef <- coefs %>% filter(term == "gdp_per_capita")
health_coef <- coefs %>% filter(term == "health_exp_pct")
r2 <- round(summary(report_model)$r.squared * 100, 1)
```

The model explains `r r2`% of the variance in life expectancy.

- **GDP per capita**: For each additional dollar in GDP per capita, life
  expectancy changes by `r round(gdp_coef$estimate, 4)` years (p =
  `r round(gdp_coef$p.value, 3)`).
- **Health expenditure**: For each 1 percentage point increase in health
  spending as a share of GDP, life expectancy changes by
  `r round(health_coef$estimate, 2)` years (p =
  `r round(health_coef$p.value, 3)`).

These results suggest that both economic development and investment in health
systems play a role in determining population health outcomes across African
countries.

---

# Summary

Today we covered:

1. **Descriptive statistics** -- `summary()`, custom tables, group summaries
2. **Hypothesis tests** -- `t.test()`, `chisq.test()`, paired tests
3. **Linear regression** -- `lm()`, interpretation, residual diagnostics
4. **Multiple regression** -- adding predictors, model comparison, ANOVA
5. **Logistic regression** -- `glm()` with binomial family, odds ratios
6. **broom** -- `tidy()`, `glance()`, `augment()` for clean model output
7. **R Markdown** -- code chunks, inline code, `kable()` tables
8. **Lab** -- a complete reproducible report from data to conclusions

Congratulations on completing the **R for Statistical Analysis** workshop!

```{r session-info}
sessionInfo()
```
