% ============================================================
% GENERATIVE AI: FROM FOUNDATIONS TO PRODUCTION
% Lecture Notes (30 hours) — English version
% Compilation: xelatex -shell-escape notes.tex (2 passes)
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\setlength{\headheight}{14.5pt}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,shapes,fit,backgrounds,chains}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[python]{python3=true}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames,svgnames]{xcolor}
\definecolor{aiblue}{RGB}{59,130,246}
\definecolor{ethicspurple}{RGB}{147,51,234}
\definecolor{warnorange}{RGB}{255,140,0}
\definecolor{deepgreen}{RGB}{22,163,74}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Custom environments ---
\newtcolorbox{aitip}{
  colback=aiblue!5,colframe=aiblue!70,
  title={\faRobot\ AI tip},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{ethicsbox}{
  colback=ethicspurple!5,colframe=ethicspurple!70,
  title={\faBalanceScale\ Ethics \& Responsible AI},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{warningbox}{
  colback=warnorange!5,colframe=warnorange!70,
  title={\faExclamationTriangle\ Caution},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{exercise}{
  colback=gray!5,colframe=gray!50,
  title={Exercise},
  fonttitle=\bfseries,breakable
}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex

% --- Icons ---
\usepackage{fontawesome5}

% --- Theorem environments ---
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}{Example}[chapter]

% ============================================================
\title{
  \vspace{-2cm}
  {\Huge\bfseries Generative AI}\\[6pt]
  {\LARGE From Foundations to Production}\\[12pt]
  {\large A 30-hour practical course}\\[24pt]
  {\normalsize Ya\'{e} Ulrich Gaba}\\[4pt]
  {\small AIRINA Labs}
}
\date{2026}
\author{}

% ============================================================
\begin{document}

% --- Title page ---
\frontmatter
\maketitle

% --- Table of contents ---
\tableofcontents
\newpage

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Generative AI has moved from research curiosity to essential infrastructure in less than four years. Large language models write code, summarize research, and answer questions. Diffusion models generate images from text descriptions. Agents chain multiple AI calls together to solve complex tasks autonomously. This is not hype --- it is the new baseline for software, research, and creative work.

This course teaches you to \emph{build} with generative AI, not just use it. You will implement tokenizers, generate text with GPT-2, fine-tune small language models with LoRA, build RAG pipelines over your own documents, generate images with Stable Diffusion, and orchestrate multi-step agents. Every example runs on free resources: Google Colab with GPU, HuggingFace open-weight models, and free API tiers from Groq and Google.

\paragraph{Who this is for.} Students and practitioners who know Python and basic machine learning but have not yet worked with large language models or diffusion models.

\paragraph{What you will be able to do after this course.}
\begin{itemize}
  \item Explain how Transformers, GPT, and diffusion models work mathematically and intuitively.
  \item Generate text with controlled decoding strategies (temperature, top-k, top-p, beam search).
  \item Write effective prompts using zero-shot, few-shot, and chain-of-thought techniques.
  \item Fine-tune a language model on custom data using LoRA/QLoRA on a single GPU.
  \item Build a complete RAG application with vector search and LangChain.
  \item Generate and edit images with Stable Diffusion and ControlNet.
  \item Evaluate generative models and identify hallucinations, biases, and safety risks.
  \item Build LLM-powered agents that use tools and reason over multiple steps.
\end{itemize}

\paragraph{Prerequisites.} Python 3.10+ (comfortable with functions, classes, pip). Basic machine learning (loss functions, gradient descent, train/test split). Linear algebra basics (vectors, matrices, dot products).

\paragraph{Software.} All code runs on Google Colab (free tier with GPU). No local GPU required.

\vfill
\noindent Ya\'{e} Ulrich Gaba\\
Cotonou, 2026

% ============================================================
\mainmatter

\include{chapters/ch01_foundations}
\include{chapters/ch02_transformers}
\include{chapters/ch03_gpt_generation}
\include{chapters/ch04_prompt_engineering}
\include{chapters/ch05_finetuning}
\include{chapters/ch06_rag}
\include{chapters/ch07_diffusion}
\include{chapters/ch08_evaluation_safety}
\include{chapters/ch09_agents}
\include{chapters/ch10_capstone}

% ============================================================
\backmatter

\chapter*{Appendix A: Environment setup}
\addcontentsline{toc}{chapter}{Appendix A: Environment setup}

\section*{Option 1: Google Colab (recommended)}
Go to \url{https://colab.research.google.com}. Select \textbf{Runtime > Change runtime type > T4 GPU}. All libraries can be installed with pip directly in notebook cells.

\section*{Option 2: Local installation}
Requires an NVIDIA GPU with 8+ GB VRAM and CUDA 12+.

\section*{Core libraries}
\begin{minted}{bash}
pip install torch transformers datasets accelerate peft bitsandbytes
pip install langchain langchain-community langchain-google-genai langgraph
pip install chromadb faiss-cpu sentence-transformers
pip install diffusers controlnet-aux
pip install tiktoken rouge-score nltk groq
\end{minted}

\chapter*{Appendix B: Free API setup}
\addcontentsline{toc}{chapter}{Appendix B: Free API setup}

\begin{longtable}{p{3cm}p{5cm}p{5.5cm}}
\toprule
\textbf{Provider} & \textbf{Free tier} & \textbf{Setup} \\
\midrule
Groq & Free API, very fast inference (Llama 3, Mixtral) & \url{console.groq.com}, get API key \\
Google Gemini & Free tier (Gemini 2.0 Flash) & \url{aistudio.google.com}, get API key \\
HuggingFace & Free inference API, free model hosting & \url{huggingface.co/settings/tokens} \\
\bottomrule
\end{longtable}

\chapter*{Appendix C: Key resources}
\addcontentsline{toc}{chapter}{Appendix C: Key resources}

\begin{itemize}
  \item Vaswani et al., ``Attention Is All You Need'' (2017) --- the original Transformer paper.
  \item Radford et al., ``Language Models are Unsupervised Multitask Learners'' (2019) --- GPT-2.
  \item Ho et al., ``Denoising Diffusion Probabilistic Models'' (2020) --- DDPM.
  \item Hu et al., ``LoRA: Low-Rank Adaptation of Large Language Models'' (2021).
  \item Lewis et al., ``Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks'' (2020).
  \item Yao et al., ``ReAct: Synergizing Reasoning and Acting in Language Models'' (2023).
  \item HuggingFace NLP Course: \url{https://huggingface.co/learn/nlp-course}
  \item Andrej Karpathy, ``Let's build GPT'': \url{https://www.youtube.com/watch?v=kCc8FmEb1nY}
  \item Jay Alammar, ``The Illustrated Transformer'': \url{https://jalammar.github.io/illustrated-transformer/}
\end{itemize}

\bibliographystyle{plainnat}
\bibliography{references}

\printindex

\end{document}
