% ============================================================
% DATA PRE-PROCESSING WITH PYTHON
% Lecture Notes (30 hours) — English version
% Compilation: xelatex -shell-escape notes.tex (2 passes)
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\setlength{\headheight}{14.5pt}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,shapes,fit,backgrounds}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[python]{python3=true}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames,svgnames]{xcolor}
\definecolor{steelblue}{RGB}{70,130,180}
\definecolor{prepgreen}{RGB}{34,139,34}
\definecolor{warnorange}{RGB}{255,140,0}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Custom environments ---
\newtcolorbox{datatip}{
  colback=steelblue!5,colframe=steelblue!70,
  title={\faDatabase\ Data tip},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{preprocesstip}{
  colback=prepgreen!5,colframe=prepgreen!70,
  title={\faCogs\ Preprocessing tip},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{warningbox}{
  colback=warnorange!5,colframe=warnorange!70,
  title={\faExclamationTriangle\ Caution},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{exercise}{
  colback=gray!5,colframe=gray!50,
  title={Exercise},
  fonttitle=\bfseries,breakable
}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex

% --- Icons ---
\usepackage{fontawesome5}

% --- Theorem environments ---
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}{Example}[chapter]

% ============================================================
\title{
  \vspace{-2cm}
  {\Huge\bfseries Data Pre-processing}\\[6pt]
  {\LARGE with Python}\\[12pt]
  {\large A 30-hour practical course}\\[24pt]
  {\normalsize Ya\'{e} Ulrich Gaba}\\[4pt]
  {\small AIRINA Labs}
}
\date{2026}
\author{}

% ============================================================
\begin{document}

% --- Title page ---
\frontmatter
\maketitle

% --- Table of contents ---
\tableofcontents
\newpage

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Data preprocessing is the unglamorous but decisive phase of every data project. Studies consistently show that data scientists spend 60--80\% of their time cleaning and preparing data. Yet most courses skip over this phase in a single lecture, eager to reach the ``interesting'' parts: models, algorithms, predictions.

This course takes the opposite approach. We devote 30 hours to the art and science of turning raw, messy, real-world data into clean, analysis-ready datasets. If you learn to preprocess data well, every model you build afterward will be better.

We use Python because its ecosystem --- \texttt{pandas}, \texttt{scikit-learn}, \texttt{missingno}, \texttt{nltk} --- provides the most complete toolkit for data preprocessing available in any language. Every example uses a real, publicly available dataset with real problems: missing values, inconsistent types, outliers, mixed encodings, and messy text.

\paragraph{What this course is not.} This is not a machine learning course (though we prepare data for ML). It is not a statistics course (though we use statistical reasoning). It is a \emph{practical data preparation course} for people who want to master the most time-consuming part of the data pipeline.

\paragraph{Prerequisites.} Basic Python (variables, lists, loops, functions). Familiarity with pandas DataFrames is helpful but will be reviewed in Chapter~1.

\paragraph{Software.} All code runs in Jupyter notebooks, either locally (Anaconda) or in the cloud (Google Colab --- free, no installation required).

\vfill
\noindent Ya\'{e} Ulrich Gaba\\
Cotonou, 2026

% ============================================================
\mainmatter

\include{chapters/ch01_data_landscape}
\include{chapters/ch02_data_loading}
\include{chapters/ch03_missing_data}
\include{chapters/ch04_outliers}
\include{chapters/ch05_type_transforms}
\include{chapters/ch06_feature_engineering}
\include{chapters/ch07_text_preprocessing}
\include{chapters/ch08_temporal_data}
\include{chapters/ch09_pipelines}
\include{chapters/ch10_capstone}

% ============================================================
\backmatter

\chapter*{Appendix A: Python installation guide}
\addcontentsline{toc}{chapter}{Appendix A: Python installation guide}

\section*{Option 1: Google Colab (recommended for beginners)}
Go to \url{https://colab.research.google.com}. Sign in with a Google account. No installation needed. All libraries used in this course are pre-installed.

\section*{Option 2: Anaconda (local installation)}
Download Anaconda from \url{https://www.anaconda.com/download}. Install with default settings. Open Jupyter Notebook from Anaconda Navigator.

\section*{Required libraries}
\begin{minted}{bash}
pip install pandas numpy matplotlib seaborn scikit-learn missingno nltk spacy joblib openpyxl sqlalchemy
\end{minted}

\chapter*{Appendix B: Dataset sources}
\addcontentsline{toc}{chapter}{Appendix B: Dataset sources}

\begin{longtable}{p{4cm}p{8cm}p{3cm}}
\toprule
\textbf{Source} & \textbf{Description} & \textbf{URL} \\
\midrule
Melbourne Housing & Housing prices with missing values and mixed types & \url{kaggle.com} \\
Titanic & Passenger survival data with missing ages and cabins & \url{kaggle.com/c/titanic} \\
FIFA 21 Raw & Player attributes with messy string-encoded numbers & \url{kaggle.com} \\
Adult Census (UCI) & Income prediction with mixed categorical/numerical & \url{archive.ics.uci.edu} \\
Jena Climate & 10-minute weather measurements over several years & \url{kaggle.com} \\
Air Quality UCI & Hourly air pollutant measurements with sensor drift & \url{archive.ics.uci.edu} \\
20 Newsgroups & Text classification dataset (20 topic categories) & via \texttt{sklearn} \\
Amazon Reviews & Product reviews for sentiment and text processing & \url{kaggle.com} \\
Gapminder & Country-level socioeconomic indicators over time & \url{gapminder.org} \\
\bottomrule
\end{longtable}

\bibliographystyle{plainnat}
\bibliography{references}

\printindex

\end{document}
