\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\setlength{\headheight}{14.5pt}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{fillbetween}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,shapes,fit,backgrounds,
  chains,decorations.pathreplacing,matrix,shapes.geometric,cd}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[python]{python3=true}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable,tabularx}

% --- Colors and boxes ---
\usepackage[dvipsnames,svgnames]{xcolor}
\definecolor{steelblue}{RGB}{70,130,180}
\definecolor{deepblue}{RGB}{0,51,102}
\definecolor{deepred}{RGB}{153,0,0}
\definecolor{deepgreen}{RGB}{0,102,51}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex


% ============================================================
% Theorem environments
% ============================================================
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}[definition]{Example}
\newtheorem{exercise}{Exercise}[chapter]
\newtheorem{proposition}[definition]{Proposition}

\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}

\theoremstyle{remark}
\newtheorem{remark}[definition]{Remark}


% ============================================================
% Colored boxes
% ============================================================
\newtcolorbox{keyformulas}[1][]{colback=blue!5,colframe=NavyBlue,
  title=\textbf{Key Formulas},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{warning}[1][]{colback=red!5,colframe=BrickRed,
  title=\textbf{Warning},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{bestpractice}[1][]{colback=green!5,colframe=ForestGreen,
  title=\textbf{Best Practice},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{codeblock}[1][]{colback=gray!5,colframe=gray!80,
  title=\textbf{Code},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{codeoutput}[1][]{colback=Apricot!5,colframe=Apricot!80!black,
  title=\textbf{Output},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{intuition}[1][]{colback=yellow!5,colframe=Goldenrod,
  title=\textbf{Intuition},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{algorithme}[1][]{colback=Cyan!5,colframe=MidnightBlue,
  title=\textbf{Algorithm},fonttitle=\bfseries,breakable,#1}
\newtcolorbox{antipattern}[1][]{colback=red!3,colframe=BrickRed!70,
  title=\textbf{Anti-pattern},fonttitle=\bfseries,breakable,#1}


% ============================================================
% Math commands
% ============================================================
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Bias}{\mathrm{Bias}}
\newcommand{\sign}{\mathrm{sign}}
\newcommand{\KL}{\mathrm{KL}}
\newcommand{\tr}{\mathrm{tr}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ind}{\mathbb{1}}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\inner}[2]{\langle #1, #2 \rangle}
\newcommand{\pkg}[1]{\texttt{#1}}
\newcommand{\cli}[1]{\texttt{#1}}
\newcommand{\file}[1]{\texttt{#1}}


% ============================================================
\newtcolorbox{attention}[1][]{colback=red!5,colframe=BrickRed,
  title=\textbf{Attention},fonttitle=\bfseries,breakable,#1}
\newcommand{\indep}{\perp\!\!\!\perp}
\newcommand{\og}{\guillemotleft{}~}
\newcommand{\fg}{~\guillemotright{}}
\begin{document}

\begin{titlepage}
\begin{tikzpicture}[remember picture, overlay]
  \fill[left color=blue!15!white, right color=blue!5!white]
    (current page.south west) rectangle (current page.north east);
  \fill[color=orange!70!yellow]
    ([yshift=-2cm]current page.north west) rectangle ([yshift=-2.3cm]current page.north east);
  \fill[color=blue!80!black, rounded corners=8pt]
    ([xshift=3cm, yshift=-4cm]current page.north west) rectangle
    ([xshift=-3cm, yshift=-10cm]current page.north east);
  \node[anchor=center, text=white, font=\Huge\bfseries]
    at ([yshift=-6cm]current page.north) {Fixed Point Theory};
  \node[anchor=center, text=orange!80!yellow, font=\Large]
    at ([yshift=-7.5cm]current page.north) {Lecture Notes};
  \node[anchor=center, text=white!80, font=\large]
    at ([yshift=-8.8cm]current page.north) {M1--PhD --- 2025--2026};
  \node[anchor=center, text=blue!80!black, font=\Large\itshape]
    at ([yshift=-12cm]current page.north) {Ya\'e Ulrich Gaba};
  \draw[orange!70!yellow, line width=2pt]
    ([xshift=5cm, yshift=-13.5cm]current page.north west) --
    ([xshift=-5cm, yshift=-13.5cm]current page.north east);
  \node[anchor=center, text width=12cm, align=center, text=blue!60!black, font=\itshape]
    at ([yshift=-16cm]current page.north) {``Every continuous function from a closed disk to itself has at least one fixed point.'' --- L.E.J. Brouwer};
  \node[anchor=south, text=gray, font=\small]
    at ([yshift=1.5cm]current page.south) {\today};
\end{tikzpicture}
\end{titlepage}

\frontmatter
\tableofcontents
\mainmatter

\input{chapters/preface}
\input{chapters/ch01_introduction}
\input{chapters/ch02_banach}
\input{chapters/ch03_extensions}
\input{chapters/ch04_generalized_spaces}
\input{chapters/ch05_brouwer}
\input{chapters/ch06_schauder}
\input{chapters/ch07_kakutani}
\input{chapters/ch08_ordered}
\input{chapters/ch09_applications}
\input{chapters/ch10_markov_kakutani}
\input{chapters/ch11_random}

\backmatter
\begin{thebibliography}{99}
\bibitem{granas} Granas, A. and Dugundji, J., \emph{Fixed Point Theory}, Springer, 2003.
\bibitem{granas2003} Granas, A. and Dugundji, J., \emph{Fixed Point Theory}, Springer, 2003.
\bibitem{smart} Smart, D.R., \emph{Fixed Point Theorems}, Cambridge, 1974.
\bibitem{smart1974} Smart, D.R., \emph{Fixed Point Theorems}, Cambridge, 1974.
\bibitem{kirk} Kirk, W.A. and Shahzad, N., \emph{Fixed Point Theory in Distance Spaces}, Springer, 2014.
\bibitem{berinde} Berinde, V., \emph{Iterative Approximation of Fixed Points}, Springer, 2007.
\bibitem{zeidler} Zeidler, E., \emph{Nonlinear Functional Analysis and Its Applications I}, Springer, 1986.
\bibitem{zeidler1986} Zeidler, E., \emph{Nonlinear Functional Analysis and Its Applications I}, Springer, 1986.
\bibitem{border} Border, K.C., \emph{Fixed Point Theorems with Applications to Economics}, Cambridge, 1985.
\bibitem{goebel1990} Goebel, K. and Kirk, W.A., \emph{Topics in Metric Fixed Point Theory}, Cambridge, 1990.
\bibitem{agarwal2001} Agarwal, R.P., Meehan, M. and O'Regan, D., \emph{Fixed Point Theory and Applications}, Cambridge, 2001.
\end{thebibliography}
\printindex
\end{document}