% ============================================================
% JULIA PROGRAMMING
% Lecture Notes (30 hours) — English version
% Compilation: xelatex -shell-escape notes.tex (2 passes)
% ============================================================
\documentclass[12pt,a4paper,twoside]{book}

% --- Encoding and language (XeLaTeX) ---
\usepackage{fontspec}
\usepackage{polyglossia}
\setmainlanguage{english}

% --- Page layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\setlength{\headheight}{14.5pt}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}

% --- Graphics ---
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,arrows.meta,calc,positioning,shapes,fit,backgrounds}

% --- Code ---
\usepackage{minted}
\setminted{fontsize=\small,frame=lines,framesep=2mm,breaklines=true}
\setminted[julia]{style=colorful}

% --- Tables ---
\usepackage{booktabs,array,multirow,longtable}

% --- Colors and boxes ---
\usepackage[dvipsnames,svgnames]{xcolor}
\definecolor{juliapurple}{RGB}{149,88,178}
\definecolor{juliared}{RGB}{203,60,51}
\definecolor{juliagreen}{RGB}{56,152,38}
\definecolor{juliablue}{RGB}{64,99,216}
\definecolor{warnorange}{RGB}{255,140,0}
\definecolor{perfgreen}{RGB}{0,128,64}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,breakable,skins}

% --- Custom environments ---
\newtcolorbox{juliatip}{
  colback=juliapurple!5,colframe=juliapurple!70,
  title={\faLightbulb\ Julia tip},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{performancetip}{
  colback=perfgreen!5,colframe=perfgreen!70,
  title={\faRocket\ Performance tip},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{warningbox}{
  colback=warnorange!5,colframe=warnorange!70,
  title={\faExclamationTriangle\ Caution},
  fonttitle=\bfseries,breakable
}
\newtcolorbox{exercise}{
  colback=gray!5,colframe=gray!50,
  title={Exercise},
  fonttitle=\bfseries,breakable
}

% --- Links ---
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=NavyBlue,citecolor=ForestGreen,urlcolor=Maroon}

% --- Bibliography and index ---
\usepackage[numbers]{natbib}
\usepackage{makeidx}
\makeindex

% --- Icons ---
\usepackage{fontawesome5}

% --- Theorem environments ---
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}{Example}[chapter]

% ============================================================
\title{
  \vspace{-2cm}
  {\Huge\bfseries Julia Programming}\\[6pt]
  {\LARGE From Fundamentals to Scientific Computing}\\[12pt]
  {\large A 30-hour practical course}\\[24pt]
  \begin{tikzpicture}
    \node[circle,fill=juliared,minimum size=0.8cm] at (0,0) {};
    \node[circle,fill=juliagreen,minimum size=0.8cm] at (0.9,0) {};
    \node[circle,fill=juliapurple,minimum size=0.8cm] at (0.45,0.78) {};
  \end{tikzpicture}\\[12pt]
  {\normalsize Ya\'{e} Ulrich Gaba}\\[4pt]
  {\small AIRINA Labs}
}
\date{2026}
\author{}

% ============================================================
\begin{document}

% --- Title page ---
\frontmatter
\maketitle

% --- Table of contents ---
\tableofcontents
\newpage

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Julia is a programming language designed for people who need both high-level expressiveness and high performance. It was created at MIT in 2012 by researchers who wanted ``the speed of C, the dynamism of Ruby, the mathematical notation of Matlab, and the generality of Python.'' A decade later, Julia has become a serious tool for scientific computing, machine learning, and data science.

This course is for anyone who already knows basic programming concepts in some language --- Python, R, Matlab, C --- and wants to learn Julia. You do not need to be an expert programmer. You need to be comfortable with the idea of variables, loops, functions, and arrays.

We cover Julia from the ground up: types, multiple dispatch, arrays, data wrangling, visualization, performance tuning, differential equations, optimization, and machine learning. Every chapter uses real datasets and real Julia code that you can run in Pluto notebooks or the Julia REPL.

\paragraph{What this course is not.} This is not a computer science theory course. It is not a numerical analysis textbook (though we do numerical computing). It is a \emph{practical programming course} for scientists, engineers, and data practitioners who want to use Julia effectively.

\paragraph{Prerequisites.} Basic programming experience in any language. Familiarity with linear algebra and calculus is helpful for Chapters 8--10 but not required for Chapters 1--7.

\paragraph{Software.} Julia 1.10 or later, VS Code with the Julia extension, and Pluto.jl for reactive notebooks. All packages are free and open source.

\vfill
\noindent Ya\'{e} Ulrich Gaba\\
Cotonou, 2026

% ============================================================
\mainmatter

\include{chapters/ch01_getting_started}
\include{chapters/ch02_types_control}
\include{chapters/ch03_functions_dispatch}
\include{chapters/ch04_arrays_linalg}
\include{chapters/ch05_dataframes}
\include{chapters/ch06_visualization}
\include{chapters/ch07_performance}
\include{chapters/ch08_scientific}
\include{chapters/ch09_ml_julia}
\include{chapters/ch10_capstone}

% ============================================================
\backmatter

\chapter*{Appendix A: Julia installation guide}
\addcontentsline{toc}{chapter}{Appendix A: Julia installation guide}

\section*{Option 1: juliaup (recommended)}
Install the Julia version manager:
\begin{minted}{bash}
# Windows (PowerShell)
winget install julia -s msstore

# macOS / Linux
curl -fsSL https://install.julialang.org | sh
\end{minted}

Then install Julia 1.10:
\begin{minted}{bash}
juliaup add 1.10
juliaup default 1.10
\end{minted}

\section*{Option 2: Direct download}
Download from \url{https://julialang.org/downloads/}. Install with default settings.

\section*{VS Code setup}
Install VS Code from \url{https://code.visualstudio.com}. In the Extensions panel, search for ``Julia'' and install the official Julia extension by julialang.

\section*{Required packages}
Open the Julia REPL and run:
\begin{minted}{julia}
using Pkg
Pkg.add([
    "Pluto", "DataFrames", "CSV", "Plots", "CairoMakie",
    "AlgebraOfGraphics", "BenchmarkTools", "DifferentialEquations",
    "Optimization", "OptimizationOptimJL", "ModelingToolkit",
    "MLJ", "Flux", "MLDatasets", "RDatasets", "IJulia"
])
\end{minted}

\chapter*{Appendix B: Key Julia resources}
\addcontentsline{toc}{chapter}{Appendix B: Key Julia resources}

\begin{longtable}{p{4cm}p{8cm}p{3cm}}
\toprule
\textbf{Resource} & \textbf{Description} & \textbf{URL} \\
\midrule
Official docs & Julia language documentation & \url{docs.julialang.org} \\
Julia Discourse & Community forum for questions & \url{discourse.julialang.org} \\
JuliaHub & Package registry and search & \url{juliahub.com} \\
Pluto.jl & Reactive notebooks for Julia & \url{github.com/fonsp/Pluto.jl} \\
SciML & Scientific ML ecosystem & \url{sciml.ai} \\
MLJ.jl & Machine learning framework & \url{alan-turing-institute.github.io/MLJ.jl} \\
Flux.jl & Deep learning library & \url{fluxml.ai} \\
Julia Academy & Free online courses & \url{juliaacademy.com} \\
\bottomrule
\end{longtable}

\bibliographystyle{plainnat}
\bibliography{references}

\printindex

\end{document}
