% notes.tex — Linear Algebra (English version)
% Compile with: xelatex -> biber -> xelatex -> xelatex
\documentclass[a4paper,11pt,twoside,openright]{book}

% Shared preamble (includes polyglossia, biblatex, etc.)
\usepackage[english]{../common/preamble}

% Bibliography file
\addbibresource{../bibliography.bib}

% ============================================================
\usepackage{tikz}
\usetikzlibrary{calc}

\begin{document}

% ============================================================
% Title page
% ============================================================
\begin{titlepage}
\begin{tikzpicture}[remember picture, overlay]
  \fill[left color=blue!15!white, right color=blue!5!white]
    (current page.south west) rectangle (current page.north east);
  \fill[color=orange!70!yellow]
    ([yshift=-2cm]current page.north west) rectangle ([yshift=-2.3cm]current page.north east);
  \fill[color=blue!80!black, rounded corners=8pt]
    ([xshift=3cm, yshift=-4cm]current page.north west) rectangle
    ([xshift=-3cm, yshift=-10cm]current page.north east);
  \node[anchor=center, text=white, font=\Huge\bfseries]
    at ([yshift=-6cm]current page.north) {Linear Algebra};
  \node[anchor=center, text=orange!80!yellow, font=\Large]
    at ([yshift=-7.5cm]current page.north) {Lecture Notes};
  \node[anchor=center, text=white!80, font=\large]
    at ([yshift=-8.8cm]current page.north) {Licence L1 --- 2025--2026};
  \node[anchor=center, text=blue!80!black, font=\Large\itshape]
    at ([yshift=-12cm]current page.north) {Ya\'e Ulrich Gaba};
  \draw[orange!70!yellow, line width=2pt]
    ([xshift=5cm, yshift=-13.5cm]current page.north west) --
    ([xshift=-5cm, yshift=-13.5cm]current page.north east);
  \node[anchor=center, text width=12cm, align=center, text=blue!60!black, font=\itshape]
    at ([yshift=-16cm]current page.north) {``The essence of mathematics lies in its freedom.''\\[4pt]--- Georg Cantor};
  \node[anchor=south, text=gray, font=\small]
    at ([yshift=1.5cm]current page.south) {\today};
\end{tikzpicture}
\end{titlepage}

\frontmatter

% ============================================================
% Table of contents
% ============================================================
\tableofcontents

% ============================================================
% Preface
% ============================================================
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}
\markboth{Preface}{Preface}

Linear algebra is one of the most fundamental and pervasive branches of modern mathematics. Its core concepts --- vector spaces, linear maps, matrices, determinants, eigenvalues --- form the foundation upon which vast areas of analysis, geometry, physics, computer science, and engineering are built.

These course notes are intended for first- and second-year undergraduate students in mathematics or engineering programs. They assume a solid background in high-school mathematics (calculus, functions, trigonometry) as well as elementary notions of set theory and logic.

\medskip

\noindent\textbf{Philosophy.}
We have striven to maintain a constant balance between \emph{algebraic rigor} and \emph{geometric intuition}. Every abstract concept is illustrated with concrete examples, figures, and applications. All proofs are complete: we believe that learning to read and write proofs is just as important as mastering computational techniques.

\medskip

\noindent\textbf{Organization.}
The notes consist of nine chapters, whose logical dependencies are shown below:

\begin{center}
\begin{tikzpicture}[
  node distance=1.2cm and 2cm,
  block/.style={rectangle, draw, rounded corners, minimum width=2.8cm,
    minimum height=0.8cm, text centered, font=\small},
  arrow/.style={-{Stealth[length=5pt]}, thick}
]
  \node[block] (ch0) {Ch.~0 Preliminaries};
  \node[block, below=of ch0] (ch1) {Ch.~1 Vector Spaces};
  \node[block, below left=1.2cm and 0.5cm of ch1] (ch2) {Ch.~2 Linear Maps};
  \node[block, below right=1.2cm and 0.5cm of ch1] (ch3) {Ch.~3 Linear Systems};
  \node[block, below=2.8cm of ch1] (ch4) {Ch.~4 Determinants};
  \node[block, below=of ch4] (ch5) {Ch.~5 Eigenvalues};
  \node[block, below left=1.2cm and 0.5cm of ch5] (ch6) {Ch.~6 Inner Products};
  \node[block, below right=1.2cm and 0.5cm of ch5] (ch8) {Ch.~8 Jordan Form};
  \node[block, below=2.8cm of ch5] (ch7) {Ch.~7 Spectral Theorem};

  \draw[arrow] (ch0) -- (ch1);
  \draw[arrow] (ch1) -- (ch2);
  \draw[arrow] (ch1) -- (ch3);
  \draw[arrow] (ch2) -- (ch4);
  \draw[arrow] (ch3) -- (ch4);
  \draw[arrow] (ch4) -- (ch5);
  \draw[arrow] (ch5) -- (ch6);
  \draw[arrow] (ch5) -- (ch8);
  \draw[arrow] (ch6) -- (ch7);
\end{tikzpicture}
\end{center}

Chapter~0 (Preliminaries) may be skimmed quickly by students already familiar with the basic concepts. Chapter~8 (Jordan Normal Form) is an advanced chapter, intended for students who wish to explore the subject in greater depth.

\medskip

\noindent\textbf{Exercises.}
Each chapter concludes with a set of exercises graded by difficulty:
\begin{itemize}
  \item \basic{} --- basic exercises (direct verification of definitions),
  \item \intermediate{} --- intermediate exercises (applications of theorems),
  \item \challenging{} --- challenging exercises (synthesis problems, open-ended questions).
\end{itemize}

\medskip

\noindent\textbf{Historical note.}
Linear algebra, as we know it today, developed gradually from the 18th to the 20th century. Major contributions include: Gauss's work on elimination (1809), Cauchy's theory of determinants (1815), the discovery of eigenvalues by Cauchy and Sylvester, Jordan's canonical forms (1870), and the axiomatic formalization of vector spaces by Peano (1888) and Banach (1920).

\vfill
\begin{flushright}
\textit{February 2026}
\end{flushright}

% ============================================================
% Notation
% ============================================================
\chapter*{Notation}
\addcontentsline{toc}{chapter}{Notation}
\markboth{Notation}{Notation}

We collect here the notation used throughout these notes. The reader is encouraged to refer back to this section as needed.

\bigskip

\begin{longtable}{@{}p{3cm}p{10cm}@{}}
\toprule
\textbf{Symbol} & \textbf{Meaning} \\
\midrule
\endhead

\multicolumn{2}{@{}l}{\textbf{Number sets}} \\[4pt]
$\N$ & Natural numbers $\{0,1,2,\ldots\}$ \\
$\Z$ & Integers \\
$\Q$ & Rational numbers \\
$\R$ & Real numbers \\
$\C$ & Complex numbers \\
$\K$ & Base field (usually $\R$ or $\C$) \\
$\F_p$ & Finite field with $p$ elements \\[6pt]

\multicolumn{2}{@{}l}{\textbf{Spaces and structures}} \\[4pt]
$\K^n$ & Space of $n$-tuples over $\K$ \\
$\Mat_{n,p}(\K)$ & Space of $n\times p$ matrices with entries in $\K$ \\
$\Mat_n(\K)$ & Square matrices of size $n\times n$ \\
$\GL_n(\K)$ & General linear group (invertible matrices) \\
$\cP_n(\K)$ & Space of polynomials of degree $\leq n$ \\
$\cL(E,F)$ & Space of linear maps from $E$ to $F$ \\
$\End(E)$ & Endomorphisms of $E$ ($=\cL(E,E)$) \\[6pt]

\multicolumn{2}{@{}l}{\textbf{Linear maps and matrices}} \\[4pt]
$\Ker f$ & Kernel of $f$ \\
$\im f$ & Image (range) of $f$ \\
$\rank(f)$, $\rank(A)$ & Rank of a linear map or matrix \\
$\Id$, $I_n$ & Identity map, $n\times n$ identity matrix \\
$\transpose{A}$ & Transpose of $A$ \\
$\hermit{A}$ & Conjugate transpose (adjoint) of $A$ \\
$\inv{A}$ & Inverse of $A$ \\
$\tr(A)$ & Trace of $A$ \\
$\det(A)$ & Determinant of $A$ \\[6pt]

\multicolumn{2}{@{}l}{\textbf{Vector spaces}} \\[4pt]
$\Span(v_1,\ldots,v_k)$ & Subspace spanned by $v_1,\ldots,v_k$ \\
$\dim E$ & Dimension of $E$ \\
$\codim F$ & Codimension of $F$ in $E$ \\
$E\oplus F$ & Direct sum \\
$E/F$ & Quotient space \\
$E^*$ & Dual of $E$ \\
$\cB=(e_1,\ldots,e_n)$ & Basis of $E$ \\[6pt]

\multicolumn{2}{@{}l}{\textbf{Inner products and norms}} \\[4pt]
$\inner{u,v}$ & Inner product of $u$ and $v$ \\
$\norm{v}$ & Norm of $v$ \\
$u\perp v$ & $u$ and $v$ are orthogonal \\
$F^\perp$ & Orthogonal complement of $F$ \\[6pt]

\multicolumn{2}{@{}l}{\textbf{Eigenvalues}} \\[4pt]
$\Spec(f)$ & Spectrum of $f$ (set of eigenvalues) \\
$E_\lambda$ & Eigenspace associated with $\lambda$ \\
$\chi_f(\lambda)$ & Characteristic polynomial of $f$ \\
$\mu_f(\lambda)$ & Minimal polynomial of $f$ \\[6pt]

\multicolumn{2}{@{}l}{\textbf{Miscellaneous}} \\[4pt]
$\deff$ & Equality by definition \\
$\delta_{ij}$ & Kronecker delta \\
$\sgn(\sigma)$ & Sign of a permutation $\sigma$ \\
$\mathfrak{S}_n$ & Symmetric group on $\{1,\ldots,n\}$ \\

\bottomrule
\end{longtable}

\mainmatter

% ============================================================
% Chapters
% ============================================================

\input{chapters/ch0-preliminaries}
\input{chapters/ch1-vector-spaces}
\input{chapters/ch2-linear-maps}
\input{chapters/ch3-linear-systems}
\input{chapters/ch4-determinants}
\input{chapters/ch5-eigenvalues}
\input{chapters/ch6-inner-products}
\input{chapters/ch7-spectral-theorem}
\input{chapters/ch8-jordan}

% ============================================================
% Back matter
% ============================================================
\backmatter

\printbibliography[heading=bibintoc,title={Bibliography}]

\printindex

\end{document}
